---
title: 'Stress and Travel'
author: 'Niloofar Nalaee'
#date: 'May 20xx'
university: 'McMaster University'
faculty: 'Faculty of Science'
department: 'School of Earth, Environment and Society'
subject: 'Geography'
supervisor: 'Antonio Paez'
bdegree: 'B.Sc.' # Bachellor's Degree
mdegree: ''   # Master's Degree (if any)
# This is the degree you are aiming for with this thesis
degree: 'M.Sc.'

knit: bookdown::render_book
site: bookdown::bookdown_site

# Remove the hashtag to specify which version of output you would like.
# Can only choose one at a time.
output:
  thesisdown::thesis_pdf: default 
  keep_tex: true
#  thesisdown::thesis_gitbook: default         
#  thesisdown::thesis_word: default
#  thesisdown::thesis_epub: default

# If you are creating a PDF you'll need to write your preliminary content 
# (e.g., abstract, acknowledgements) below or use code similar to line 25-26 
# for the .RMD files. If you are NOT producing a PDF, delete or silence
# lines 31-56 in this YAML header.
lay_abstract: |
  The lay abstract must be 150 words or less.
  
  It must explain the key goals and contributions of the thesis in lay terms that are accessible to the general public. 
abstract: |
  This is the abstract.
  
  I can write a really long abstract.
# Note that a tab is needed on the line after the `|`. Subsequent paragraphs need to be indented too.
acknowledgements: |
  I want to thank a few people.
  
  This includes my friends.
dedication: |
  I wish to dedicate this to my awesome brother. 
declarations: |
  I did most of the research.
  
  Also the writting. 
  
  Sometimes I cried.
  
  But mostly I had fun.
quotation: 'You may hate gravity, but gravity does not care'
# Author of the quotation
qauthor: 'Albert Einstein'
# Specify the location where the bibliography file is found below
bibliography: bib/thesis.bib
# Download your specific csl file and refer to it in the line below.
csl: csl/apa.csl
lot: true
lof: true
---

<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of 
metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete the section entirely, or silence them (add # before each line). 

If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.

If you receive a duplicate label error after knitting, make sure to delete the index.Rmd file and then knit again.
-->

```{r include_packages, include = FALSE}
# This chunk ensures that the macdown package is
# installed and loaded. This macdown package includes
# the template files for the thesis.
if(!require(remotes))
  install.packages("remotes", repos = "http://cran.rstudio.com")
if(!require(macdown))
  remotes::install_github("paezha/macdown")
library(macdown)
```

<!--
The acknowledgements, preface, dedication, and abstract are added into the PDF
version automatically by inputting them in the YAML at the top of this file.
-->

<!-- The {.unnumbered} option here means that the introduction will be 
"Chapter 0." You can also use {-} for no numbers on chapters.
-->

# Preface {.unnumbered}

Welcome to the _R Markdown_ thesis template. This template is based on (and in many places copied directly from) the Reed College LaTeX template, but hopefully it will provide a nicer interface for those that have never used TeX or LaTeX before.  Using _R Markdown_ will also allow you to easily keep track of your analyses in **R** chunks of code, with the resulting plots and output included as well.  The hope is this _R Markdown_ template gets you in the habit of doing reproducible research, which benefits you long-term as a researcher, but also will greatly help anyone that is trying to reproduce or build onto your results down the road.

Hopefully, you won't have much of a learning period to go through and you will reap the benefits of a nicely formatted thesis.  The use of LaTeX in combination with _Markdown_ is more consistent than the output of a word processor, much less prone to corruption or crashing, and the resulting file is smaller than a Word file. While you may have never had problems using Word in the past, your thesis is likely going to be about twice as large and complex as anything you've written before, taxing Word's capabilities.  After working with _Markdown_ and **R** together for a few weeks, we are confident this will be your reporting style of choice going forward.

<!-- 
If you're still on the fence about using _R Markdown_, check out the resource
for newbies available at <https://rbasics.netlify.com>. 
-->

**Why use it?**

_R Markdown_ creates a simple and straightforward way to interface with the beauty of LaTeX.  Packages have been written in **R** to work directly with LaTeX to produce nicely formatting tables and paragraphs. In addition to creating a user friendly interface to LaTeX, _R Markdown_ also allows you to read in your data, to analyze it and to visualize it using **R** functions, and also to provide the documentation and commentary on the results of your project.  Further, it allows for **R** results to be passed inline to the commentary of your results.  You'll see more on this later.  

<!-- 
Having your code and commentary all together in one place has a plethora of 
benefits!
-->

**Who should use it?**

Anyone who needs to use data analysis, math, tables, a lot of figures, complex cross-references, or who just cares about the final appearance of their document should use _R Markdown_. Of particular use should be anyone in the sciences, but the user-friendly nature of _Markdown_ and its ability to keep track of and easily include figures, automatically generate a table of contents, index, references, table of figures, etc. should make it of great benefit to nearly anyone writing a thesis project.

**For additional help with bookdown** 

Please visit [the free online bookdown reference guide](https://bookdown.org/yihui/bookdown/).

<!--

BEGINNING HERE ADD THE DOCUMENTS THAT CONFORM THE MAIN BODY OF THE THESIS

-->

```{r eval=!knitr::is_latex_output(), child=here::here("01-chap1.Rmd")}

```

```{r eval=!knitr::is_latex_output(), child=here::here("02-chap2.Rmd")}

```

```{r eval=!knitr::is_latex_output(), child=here::here("03-conclusion.Rmd")}

```

```{r eval=!knitr::is_latex_output(), child=here::here("99-references.Rmd")}

```

<!--chapter:end:index.Rmd-->

<!--
This is for including Chapter 1.  Notice that it's also good practice to name your chunk.  This will help you debug potential issues as you knit.  The chunk above is called intro and the one below is called chapter1.  Feel free to change the name of the Rmd file as you wish, but don't forget to change it here from chap1.Rmd.
-->

<!--
The {#rmd-basics} text after the chapter declaration will allow us to link throughout the document back to the beginning of Chapter 1.  These labels will automatically be generated (if not specified) by changing the spaces to hyphens and capital letters to lowercase.  Look for the reference to this label at the beginning of Chapter 2.
-->
<!--
```{r install-data-package, include = FALSE}
#| label: install-data-package
#| #| include: false
# Run only once if needed to install data package `sobiEquity`
# if (!require("bSantiago", character.only = TRUE)) {
#       remotes::install_github("paezha/bSantiago")
#   }
```
--->

```{r load-packages, include=FALSE}
library(bSantiago) # A data package with the results of a travel and well-being survey conducted in Santiago in 2016
library(dplyr) # A Grammar of Data Manipulation
library(skimr) # Compact and Flexible Summaries of Data
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(ggplot2)
library(reshape2)
library(plyr)
library(readxl)
```

```{r load-data, include=FALSE}
data("Santiago_BE")
data("Santiago_FE")
data("Santiago_H")
data("Santiago_IC")
data("Santiago_IS")
data("Santiago_NS")
data("Santiago_RPD")
data("Santiago_SI")
data("Santiago_TW")
```

# Abstract

Large-scale travel surveys are invaluable sources of information to understand travel behavior and other aspects of the urban experience, such as residential context. By their nature, they often shy away from overloading respondents with additional questions. The data presented in this article result from an effort to purposefully collect data on various aspects of the experience of living and moving in a major city in the Global South. The data set contains essential socio-economic and demographic information about the respondents, as well as their built environment and behaviors commuting to work. In addition, the survey (conducted between DATE-DATE, 2016) includes information about the respondents' feelings and emotions in relation to their commuting experience, the social experience of a variety of transportation modes, various self-assessed health questions, patterns of use of information and telecommunication technologies, and questions about sustainability and the environment. The survey was adopted a quota-sampling method that used the Pre-Census of 2012 as a frame, and in total includes 451 validated questionnaires.

<!-- what do we mean by validated -->

<!--

The data set presented in this article would be advantageous to give intuition about the experience of daily travel and its psychological impact on travelers, ranging from positive feelings of enjoyment in some to the sensation of stress in many others. This dataset particularly examines the feelings of stress by users of active and motorized modes of transportation. Furthermore, it also investigates the importance that travelers attach to their feelings of stress. This allows us to explore the concept of “limited horizons”, the normalization of subpar experiences by those less able to adapt. Moreover, it is about the functioning, freedom and an appropriate environment for human beings to be flourished and refers to the actual ability that people have and can do [@walker2005amartya]. According to the capability approach, functioning is the achievement while the capability is the ability to achieve[@rajapakse2016amartya]. This data provides information about commuters in terms of their individual characteristics, health-related factors while using transportation modes, feelings and emotions towards different modes of travel, own decisions, social interaction, attitudes towards nature and sustainability, shifts between modes, attitudes towards built environment, traveling to work. Data for the research are drawn from a survey conducted in Santiago, Chile, based on a quota-sampling method based on the information from Pre-Census of 2012, and in total, 451 persons validly completed the survey. Describe the statistics and conclude suggest that
-->

## Key words 

- Transportation
- Commuting to work
- Built environment
- Well-being
- Equity
- Santiago
- Chile

## Specifications Table

See @tbl-specifications for details of the data set.

| Items | Explanation |
|:--------|:-----|
| Subject area      | Transportation, Geography, Public Health and Health Policy, Urban development   |
| More specific subject area     | Transport inequalities, Stress and limited horizons, Travel behavior, Global South  |
| Type of data       | `R` Data Package    |
| How data was acquired | The survey was conducted using a (pen-and-paper??) questionnaire. The instrument contains some quantitative variables regarding the individual characteristics of respondents and mostly 5-point Likert scale responses in the rest of the questionnaire|
| Data format | Thematic tables and documentation in native `R` format. The thematic tables can be linked by means of a common `ID` field | 
| Parameters for data collection | The survey was collected using a quota-sampling method based on the information from Pre-Census of 2012, and in total, 451 persons validly completed the survey and face-to-face in Santiago, Chile in 2016. The survey collected information on a wide range of travel-related issues (socio-demographics, health-related, perceptions and travel behavior, travel choices and planning, social interaction factors, built environment, among others) |
| Description of data collection | Data was acquired through the 5-Likert scale questionnaire regarding most sections of the questionnaire, using a face-to-face and quota-sampling method for individual characteristics  | 
| Data source location | Santiago, Chile | 
| Data accessibility | https://paezha.github.io/bSantiago/ | 

: Specifications table {#tbl-specifications}

## Value of the data 

A large body of the data has made inroads investigating psychological impact on travelers ranging from positive feelings of enjoyment in some to the sensation of stress in many others that can affect the effectiveness of policy measure (in the case of positive feelings) and are known to affect health outcomes (in the case of stress).This would be interesting for those with transport policies concerns as public feelings like happiness and satisfaction rating of travel mode switching has been studied previously [@abou2012happiness].
Data set contribute to psychological impact on travelers both active and motorized modes of transportation to examines not only the feeling of stress, but also how these effects are experienced by travelers and investigates the importance that travelers attach to their feelings of stress which makes it valuable for researchers who focused on public sector development and health-related policies.
It allows us to realize the concept of “limited horizons” and focus on developing people's capability to select what they have their reason behind it. Furthermore, it would be advantageous for decision makers to evaluate their attempts to flourish well-being by expanding capabilities in further research regarding transport inequalities, index of stress, travel behavior in the region or even as a representative for other areas with similar attributes[@walker2005amartya].
The dataset provides a wide range of travel-related issues such as socio-demographics, health-related, perceptions and travel behavior, travel choices and planning, social interaction factors, built environment, among others.  

## Data
 
The data package described in this paper contains 9 data objects in tabular format, which are listed in @tbl-list-of-tables. Each object corresponds to a theme and includes information about the individual characteristics, travel-to-work related information, and perceptions of the built environment. Furthermore, other themes include social interactions by mode of transportation, mode shifting and use of information technologies, a number of self-assessed health responses, feelings and emotions as they relate to the commute experience, attitudes with respect to how well transportation systems satisfy the needs of respondents, and attitudes about nature and sustainability. As seen in @tbl-list-of-tables, the number of variables in these tables ranges from 5 to 24. The tables share a common unique identifier (`ID`) that can be used to link several tables in a consistent way for multi-theme analysis of the data.

| Table | Theme |
|:--------|:-----|
| Santiago_IC      | A table with `r ncol(Santiago_IC)` variables that describe the individual characteristics of respondents   |
| Santiago_TW     | A table with `r ncol(Santiago_TW)` variables regarding the commuting behavior of respondents  |
| Santiago_SI     | A table with `r ncol(Santiago_SI)` variables about social interaction by mode of transportation |
| Santiago_IS     | A table with `r ncol(Santiago_IS)` variables related to mode-shifting and use of information technologies  |
| Santiago_BE  | A table with `r ncol(Santiago_BE)` variables about built environment at the place of residence of respondents  |
| Santiago_H     | A table with `r ncol(Santiago_H)` variables about health attributes of respondents  |
 Santiago_FE     | A table with `r ncol(Santiago_FE)` variables with information about the feelings and emotions of respondents  |
| Santiago_RPD     | A table with `r ncol(Santiago_RPD)` variables with information about attitudes with respect to the effectiveness of the transportation system  |
| Santiago_NS     | A table with `r ncol(Santiago_NS)` attitudinal variables that refer to nature and sustainability aspects of city  |

: List of tables in data package {#tbl-list-of-tables}

The theme of table `Santiago_IC` is the individual characteristics of respondents and includes several essential socio-economic and demographic attributes. @tbl-individual-characteristics-descriptives) presents a summary of these variables, including information about the neighborhood of residence of respondents. All respondents were successfully geolocated to the level of Communes (large administrative partitions equivalent to municipalities) and Macro Zones (larger aggregations of communes used for descriptive purposes). The street address information is less complete. The place of work (geocoded at the level of Communes) is missing only in a small number of cases (complete rate 0.96) and few respondents report a _second_ place of work/studies where they travel (`r01_WORK_COM_2`).

As seen in the table, most respondents' information is quite complete. Only 16 respondents reported facing some disability, and of these only 11 completed the section asking them about the type (`r0O_TYPE_DIS`). In terms of gender, the sample includes approximately 15.7\% more women than men. 

**BEA:** PLEASE COMMENT ON THE DISTRIBUTION OF AGES, EDUCATION, OCCUPATION, AND INCOME; ALSO, IS THE VARIABLE MONTHLY OR ANNUAL INCOME? HOW WHERE THE LEVELS/CUTS FOR INCOME SELECTED? ARE THE UNITS IN PESOS OR 1,000 PESOS?

**BEA:** r0P_MODE_1 IS PRIMARY MODE? AND r0Q_MODE_2 AND r0R_MODE_3 ARE SUBSEQUENT MODES FOR THE SAME TRIP? OR MODES USED IN TERMS OF DECREASING FREQUENCY?

Other variables regarding the respondents include their height (cm) and weight (kg), and the household context in terms of number of adults (between 1 and 7) and minors in the family (75% of respondents report no children, and of the remainder the number of children varies between 1 and 6).

These variables provide insights about different segments of the population. In terms of place of residence (**BEA:** ARE ALL COMMUNES URBAN? IS THERE AN URBAN-RURAL INTERFACE?). 

<!-- there should be a more specified classification than in urban, semi-rural and rural areas regarding different transportation options that they might have[@bergstad2011affective]. Also this part would be advantageous for business developments to figure out marketing and apply strategies for future patterns based on consumer demands.-->

```{r Summary-individual-characteristics}
#| tbl-cap: Variables regarding individual characteristics of respondents
#| tbl-cap-location: top
#| label: tbl-individual-characteristics-descriptives
#| echo: false

#skim(Santiago_IC)
skim_IC <- Santiago_IC |>
  select(-ID) |>
  skim()

skim_IC$factor.top_counts[9] <- skim_IC$factor.top_counts[9] |>
  str_replace_all(c("35t" = "35 to 54", 
                    "18t" = "18 to 24",
                    "25t" = "25 to 34",
                    "55t" = "55 to 64"))

skim_IC$factor.top_counts[12] <- skim_IC$factor.top_counts[12] |>
  str_replace_all(c("977" = "977 to 1,550", 
                    "238" = "2,380 or more",
                    "Les" = "Less than 423",
                    "155" = "1,550 to 2,380"))

skim_IC
```

```{r}

par(mfrow=c(1,2))

# Age and Gender Bar Graph
#SantiagoSurvey2016 <- read_excel(choose.files())



# ggplot(SantiagoSurvey2016,aes(x=factor(r0D_EDAD), fill=factor(r0C_GENERO)))+
#   geom_bar(position="dodge")+
#   scale_x_discrete("Age Groups", 
#                    labels=c("A"="Less18","B"="18-24","C"="25-34",
#                             "D"="35-54","E"="55-64","F"="65+"))+
#   scale_fill_discrete(name="Gender",
#                       breaks=c("A", "B"),
#                       labels=c("Female", "Male"))+
#   ggtitle("Age Groups by Gender")+
#   ylab('No. of People')


ggplot(Santiago_IC,
       aes(x = factor(r0D_AGE), fill=r0C_SEX))+
  geom_bar(position="dodge")+
  scale_x_discrete("Age Groups", 
                   labels=c("A"="Less18","B"="18-24","C"="25-34",
                            "D"="35-54","E"="55-64","F"="65+"))+
  scale_fill_discrete(name="Gender",
                      breaks=c("A", "B"),
                      labels=c("Female", "Male"))+
  ggtitle("Age Groups by Gender")+
  ylab('No. of People')

# Education bar graph
qplot(Santiago_IC$r0G_EDUCATION, 
      geom='bar', 
      fill=Santiago_IC$r0C_SEX)+
  coord_flip()+
  scale_x_discrete("Education Levels",
                   labels=c("A"="ELEMENTARY","B"="SECUNDARY",
                            "C"="PROFESS_TECH","D"="COLLEGE",
                            "E"="POSTGRAD"))+
  scale_fill_discrete(name="Gender",
                      breaks=c("1", "2"),
                      labels=c("Male", "Female"))+
  ggtitle("Education Levels by Gender")+ylab('No. of People')
```

```{r}
#Employment Pie Chart
# SantiagoSurvey2016$r0H_OCUPACION <- reorder(SantiagoSurvey2016$r0H_OCUPACION, X = SantiagoSurvey2016$r0H_OCUPACION, 
#                             FUN = function(x) -length(x))
# 
# at <- nrow(SantiagoSurvey2016) - as.numeric(cumsum(sort(table(SantiagoSurvey2016$r0H_OCUPACION)))-
#                                   0.5*sort(table(SantiagoSurvey2016$r0H_OCUPACION)))
# 
# label=paste0(round(sort(table(SantiagoSurvey2016$r0H_OCUPACION))/
#                      sum(table(SantiagoSurvey2016$r0H_OCUPACION)),2) * 100,"%")

# ggplot(Santiago_IC,
#        aes(fill = r0H_OCCUPATION))

#p <- 
ggplot(Santiago_IC,
       aes(x = r0H_OCCUPATION,
                                   fill = r0H_OCCUPATION)) +
  geom_bar() +
  # scale_fill_discrete(name="Occupation",
  #                     breaks=c("A","B","C", "D", "E", "F", "G", "H"),
  #                     labels=c("FULL_TIME", "PART_TIME","SELF_EMPLOYED", "UNEMPLOYED", "HOME_TAKER", "STUDENT", "STUDENT_WORK", "RETIRED"))+
  #annotate(geom = "text", aes(x = r0H_OCUPATION), y = at, x = 1, label = label,size=3)+
  #coord_polar(theta="y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  ggtitle("Occupation Type")
#print(p)
```

```{r}
# df <- data.frame(matrix(ncol = 2, nrow = 6)) 
# colnames(df)<-c('Income','Frequency')
# df$r0J_INGRESO<-c('Less423','423to639', '639to977','977to1550','1550to2380',
#            '2380more')
# 
# for (i in 1:6){df[i,2]<-sum(SantiagoSurvey2016[,77+i])}
# 
# df$r0J_INGRESO<-factor(df$r0J_INGRESO, levels=df$r0J_INGRESO) #to prevent reordering of bars
# 
# 
# ggplot(df,aes(x=r0J_INGRESO, y=Frequency), fill=df$r0J_INGRESO)+
#   geom_bar(stat="identity")+
#   scale_x_discrete("Personal Income Groups")+
#   coord_flip()+
#   ggtitle('Personal Income Distribution')+
#   ylab('No. of People')
```


The theme of the next table, `Santiago_TW`, deals with commuting and work variables (see @tbl-Travel-work-Descriptives) and consists of seven ordinal categorical variables (factors). Variable `r8A_ACCESSJOB` refers to the impact that respondents feel the transportation network has affected their chances of having better jobs. The most common responses were `SOME IMPACT` and `NO IMPACT`, but we see that approximately 14.2\% of respondents feel that the network has had a major impact. This variable tracks to some extent with the responses to job opportunities in the commune of residence (`r8B_JOBOPP`), suggesting a possible correlation between local opportunities and the impact of the transportation network on job outcomes. When asked about their ideal level of accessibility in the commune of the residence (`r8C_ACC_COM`), a majority respondents opt for excellent and very good (**BEA:** I AM A LITTLE UNCLEAR ABOUT THE MEANING OF THE QUESTION THAT PRODUCES THIS VARIABLE).

In terms of the level of satisfaction with their current job, we see that almost 60\% of respondents are at least highly satisfied. We also see that long commutes are frequent in this sample, with about one third or respondents spending 1 h or more travelling (**BEA:** ONE WAY OR TWO WAYS?) and about one quarter of respondents spending between 40 minutes and one hour in their daily commute. This distribution is noteworthy because time spent commuting has been recognized as a factor that can affect physical and mental health and well-being in particular in association with motorized transportation [@brutus2017cycling].

The most common time of the day for commuting is between 7 am and 9 am, but there are also 171 missing responses in this column, so not much can be read from it. Finally, we note that many people spend 35,000-75,000 (**BEA:** WHAT ARE THE UNITS OF THIS? IS IT 35 PESOS OR 35,000 PESOS) monthly on their transportation expenditure. (**BEA:** MAYBE COMMENT ON THE TYPICAL RATIO OF TRANSPORT EXPENDITURE TO INCOME).

```{r summary-travel-to-work}
#| tbl-cap: Variables regarding the commuting behavior of respondents
#| tbl-cap-location: top
#| label: tbl-Travel-work-Descriptives
#| echo: false
#
#skim(Santiago_TW)
skim_STW <- Santiago_TW |>
  select(-ID) |>
  skim()

skim_STW$factor.top_counts[1] <- skim_STW$factor.top_counts[1] |>
  str_replace_all(c(SOM = "SOME IMPACT", 
                    NO = "NO IMPACT",
                    MIN = "MINOR IMPACT",
                    MAJ = "MAJOR IMPACT"))

skim_STW$factor.top_counts[2] <- skim_STW$factor.top_counts[2] |>
  str_replace_all(c(GOO = "GOOD",
                    POO = "POOR",
                    FAI = "FAIR",
                    VER = "VERY GOOD"))

skim_STW$factor.top_counts[3] <- skim_STW$factor.top_counts[3] |>
  str_replace_all(c(EXC = "EXCELLENT",
                    GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR"))

skim_STW$factor.top_counts[4] <- skim_STW$factor.top_counts[4] |>
  str_replace_all(c(HIG = "HIGH SATISF",
                    VHI = "VERY HIGH SATISF",
                    MED = "MEDIUM SATISF",
                    LOW = "LOW SATISF"))

skim_STW$factor.top_counts[5] <- skim_STW$factor.top_counts[5] |>
  str_replace_all(c("1h" = "1h and more",
                    "40-" = "40-60 min",
                    "20-" = "20-40 min",
                    "0-2" = "0-20 min"))

skim_STW$factor.top_counts[6] <- skim_STW$factor.top_counts[6] |>
  str_replace_all(c("9:0" = "9:00 - 13:00",
                    "7:0" = "7:00 - 9:00",
                    "Oth" = "Others",
                    "18:" = "18:00 - 21:00"))

skim_STW$factor.top_counts[7] <- skim_STW$factor.top_counts[7] |>
  str_replace_all(c("75." = "75,000-125,000",
                    "35." = "35,000-75,000",
                    "Les" = "LESS THAN 35,000",
                    "Mor" = "MORE THAN 125,000"))
  
skim_STW
```

Table `Santiago_SI` includes five variables that capture various aspects of social interaction while commuting (see @tbl-Social-Interaction-Descriptives). Social interaction is a topic of interest for mode-related choices given earlier evidence that for some commuters privacy is an important consideration and/or a way to manage social stressors [@gardner2007drives; @paez2010enjoyment; @lowe2016conceptual]. When asked to rate the level of interaction with other people during their usual trips, respondents  In terms of the level of interaction people have with others during their usual trips, most of them presume a good level and they moderately feel it is important when they consider the presence of other people during their usual travels (`r4A_INTERACC`) a plurality of responses are "POOR" or "FAIR" (187) and only 73, that is 16.2\% of respondents, rate their level of interaction as "GOOD" or "EXCELLENT" report poor or fair (there are 13 missing responses). With respect to the presence of other people during their travels, the responses tend to be somewhat more ambivalent, and the difference between those for whom this is less or more important is smaller (157 responses are "NOT" or only "SLIGHTLY IMPORTANT" whereas 132 responses are "IMPORTANT" or "VERY IMPORTANT").

The next two variables in this table deal with feelings and the experience of discrimination: "have you felt discriminated against while travelling?" (`r4C_DISCRIM`) and if so, "while using which mode of transportation" (`r4D_MODE`). We can see in @tbl-Social-Interaction-Descriptives, that slightly fewer than one quarter of respondents (24.2\%) report having felt discriminated while commuting (only 4 responses are missing), and of those close to half (44\%) had that experience in public transportation (however, note that about 58\% of those who felt discriminated against did not state the mode).

```{r summary-social-interaction}
#| tbl-cap: Variables regarding social interactions of respondents
#| tbl-cap-location: top
#| label: tbl-Social-Interaction-Descriptives
#| echo: false
#
#skim(Santiago_SI)
skim_SSI <- Santiago_SI |>
  select(-ID) |>
  skim()

skim_SSI$factor.top_counts[1] <- skim_SSI$factor.top_counts[1] |>
  str_replace_all(c(GOO = "GOOD", 
                    FAI = "FAIR",
                    POO = "POOR",
                    VER = "VERY GOOD"))

skim_SSI$factor.top_counts[2] <- skim_SSI$factor.top_counts[2] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SSI$factor.top_counts[4] <- skim_SSI$factor.top_counts[4] |>
  str_replace_all(c(MET = "METRO",
                    TAX = "TAXI"))

skim_SSI
```

```{r eval=FALSE}
#| include: false
# THERE IS AN ERROR WITH n() NOT SURE WHAT IS CAUSING IT
Santiago_IC |> 
  left_join(Santiago_SI, by = "ID") |>
  select(r0C_SEX, r4C_DISCRIM, r0P_MODE1) |>
  group_by(r0C_SEX, r4C_DISCRIM, r0P_MODE1) |>
  summarize(discrim = n())
```

**BEA:** Niloofar described @tbl-ITC-Shifting-Descriptives as "telecommunication and shifting different modes" in the documentation, but I would like to ask for clarification, because it seems to me that the questions relate to _transfers_. For example, the question:

- Do you shift between transport modes on your usual trips?	

Seems to refer to transfers between public transportation modes (e.g., metro to bus), and then the next questions refer to how long waits are, and ease of transferring.

- If YES in the previous question, how is the quality of these inerchanges?	

- How would you assess your regular trips in terms of: Waiting times	How would you assess your regular trips in terms of: Time of shifts	

- How would you assess your regular trips in terms of: Travel time	What main aspect makes your shifts difficult?	

The last three questions relate to route planning tools, not ITC more generally (for instance, this is unrelated to working from home).

- Do you have access to technology tools such as smartphones and internet, with which you can view information on transportation services?	

- How do you assess the level of technological information available to see transportation alternatives? (eg smartphone applications, internet, signs)	

- How important is it for you to have access to technological information for your usual trips?

Please comment on these variables!

**THIS IS NILOOFAR'S TEXT:**

The information, telecommunications and mode shifting of respondents (see @tbl-ITC-Shifting-Descriptives) reveals numerous factorial variables using Likert scale to identify the exact level of respondents' viewpoints. We can see a considerable number of missing values in quality of changing mode of travel because of people's decision on mostly saying yes to shift between transport modes on their usual trips and for those who change the quality of these inerchanges was good. Many people assign a good level when they are asked to assess the waiting times, time of shifts and total travel time in their regular trips. Most people have access to technology tools such as smartphones and internet, with which they can view information on transportation services and they assign a good level of technological information available to see transportation alternatives (eg smartphone applications, internet, signs) and most of them assume it is very important to have access to technological information for their usual trips. 

This section will gives insights on designing a strong transportation network to facilitate people's different activities such as work, school, grocery shopping and so forth. Also experts can make users informed about traffic circumstances, road information and cost of transportation by using various modes. Following this, car-use deduction has been turned into the political agenda around the world due to the increasing negative effects of motorized modes of travel on environment and public health. Therefore, improvements in public transportation sector are needed to achieve this goal regarding previous studies that has shown participants are more likely to use bus with increased frequencies, shorter travel time, and high access to bus stops[@ettema2011satisfaction].

```{r summary-ITC-and-shifting}
#| tbl-cap: Variables regarding information and telecommunications and mode shifting of respondents
#| tbl-cap-location: top
#| label: tbl-ITC-Shifting-Descriptives
#| echo: false
#
#skim(Santiago_IS)
skim_SIS <- Santiago_IS |>
  select(-ID) |>
  skim()

skim_SIS$factor.top_counts[2] <- skim_SIS$factor.top_counts[2] |>
  str_replace_all(c(GOO = "GOOD",
                    FAI = "FAIR",
                    VER = "VERY GOOD",
                    POO = "POOR"))

skim_SIS$factor.top_counts[3] <- skim_SIS$factor.top_counts[3] |>
  str_replace_all(c(GOO = "GOOD",
                    POO = "POOR",
                    FAI = "FAIR",
                    VER = "VERY GOOD"))

skim_SIS$factor.top_counts[4] <- skim_SIS$factor.top_counts[4] |>
  str_replace_all(c(GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR",
                    POO = "POOR"))

skim_SIS$factor.top_counts[5] <- skim_SIS$factor.top_counts[5] |>
  str_replace_all(c(GOO = "GOOD",
                    POO = "POOR",
                    FAI = "FAIR",
                    VER = "VERY GOOD"))

skim_SIS$factor.top_counts[6] <- skim_SIS$factor.top_counts[6] |>
  str_replace_all(c(All = "ALL THE PREVIOUS ONES",
                    Too = "TOO LONG SHIFTS",
                    Unc = "UNCOMFORTABLE SHIFTING",
                    Bad = "BAD INFRASTRUCTURE FOR WAITING TIMES"))

skim_SIS$factor.top_counts[8] <- skim_SIS$factor.top_counts[8] |>
  str_replace_all(c(GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SIS$factor.top_counts[9] <- skim_SIS$factor.top_counts[9] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SIS
```

The theme of table `Santiago_BE` is perceptions of the built environment. The way environments are perceived, and not just their objective attributes, has been shown to correlate with travel behavior [e.g., @Jamal2020perceptions; @Loo2015transport]. Surveying perceptions is therefore a good remedy to the lack of canonical data sets in regions of the world where built environment attributes are not systematically collected. 

There are in total 22 variables in this table, and their descriptive statistics appear in @tbl-Built-Environment-Descriptives. The variables are organized in pairs: one asks about the perception of an attribute and the second the importance of that attribute to the respondent. In this way, `r7AA_AUTOSPACE` is about the perception of space for autos, and `r7BA_AUTOSPACE` is about the importance that respondents assign to this attribute. Aspects of the built environment covered by this table, in addition to space for autos, are number of parking spaces, quality of highways (**BEA:** ROADS IN GENERAL OR HIGHWAYS SPECIFICALLY?), space for pedestrians, quality of sidewalks, cleanliness of bus stops and seating areas, protection from inclement weather at bus stops, amount and quality of cycleways, and bike sharing schemes. 

The descriptive statistics suggest that most respondents have positive evaluations of space for cars and parking spaces, at the same time that they assign a high level of importance to these attributes. Respondents also show a slight tendency to assess positively the space for pedestrians and quality of sidewalks located near of their home, and these features are also regarded as very important or important by a majority of respondents. In contrast, although respondents feel strongly about the importance of facilities related to buses and cycling, the perception of these attributes tends to be poor. 
```{r summary-built-environment}
#| tbl-cap: Variables regarding the built environment at the place of residence of respondents
#| tbl-cap-location: top
#| label: tbl-Built-Environment-Descriptives
#| echo: false
#
#skim(Santiago_BE)
skim_SBE <- Santiago_BE |>
  select(-ID) |>
  skim()

skim_SBE$factor.top_counts[1] <- skim_SBE$factor.top_counts[1] |>
  str_replace_all(c(GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SBE$factor.top_counts[2] <- skim_SBE$factor.top_counts[2] |>
  str_replace_all(c(GOO = "GOOD",
                    FAI = "FAIR",
                    VER = "VERY GOOD",
                    POO = "POOR"))

skim_SBE$factor.top_counts[3] <- skim_SBE$factor.top_counts[3] |>
  str_replace_all(c(GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SBE$factor.top_counts[4] <- skim_SBE$factor.top_counts[4] |>
  str_replace_all(c(GOO = "GOOD",
                    VER = "VERY GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SBE$factor.top_counts[5] <- skim_SBE$factor.top_counts[5] |>
  str_replace_all(c(GOO = "GOOD",
                    FAI = "FAIR",
                    VER = "VERY GOOD",
                    POO = "POOR"))

skim_SBE$factor.top_counts[6] <- skim_SBE$factor.top_counts[6] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD",
                    VER = "VERY GOOD"))

skim_SBE$factor.top_counts[7] <- skim_SBE$factor.top_counts[7] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD",
                    VER = "VERY GOOD"))

skim_SBE$factor.top_counts[8] <- skim_SBE$factor.top_counts[8] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD",
                    VER = "VERY GOOD"))

skim_SBE$factor.top_counts[9] <- skim_SBE$factor.top_counts[9] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD",
                    VER = "VERY GOOD"))

skim_SBE$factor.top_counts[10] <- skim_SBE$factor.top_counts[10] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD",
                    VER = "VERY GOOD"))

skim_SBE$factor.top_counts[11] <- skim_SBE$factor.top_counts[11] |>
  str_replace_all(c(POO = "POOR",
                    GOO = "GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SBE$factor.top_counts[12] <- skim_SBE$factor.top_counts[12] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[13] <- skim_SBE$factor.top_counts[13] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[14] <- skim_SBE$factor.top_counts[14] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SBE$factor.top_counts[15] <- skim_SBE$factor.top_counts[15] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SBE$factor.top_counts[16] <- skim_SBE$factor.top_counts[16] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[17] <- skim_SBE$factor.top_counts[17] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[18] <- skim_SBE$factor.top_counts[18] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[19] <- skim_SBE$factor.top_counts[19] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[20] <- skim_SBE$factor.top_counts[20] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SBE$factor.top_counts[21] <- skim_SBE$factor.top_counts[21] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT"))


skim_SBE$factor.top_counts[22] <- skim_SBE$factor.top_counts[22] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))


skim_SBE
```

Table `Santiago_H` contains responses to twelve questions related to health. 
Health information could be useful in investigating affects of transportation policy decisions on public health. Also being aware of specific factors making modes stressful will help transportation and public health experts make commuting a safer, more enjoyable and less stressful activity. Consequently ,planners and transportation experts could mitigate the potentially serious health outcomes of a stressful commute [@legrain2015stressed]. Having such a detailed data set and understanding the relationship between health and mode choice in commuting would help them to adopt reform policies and management strategies in accordance with healthy modes of travel as active and public transportation [@mattisson2018modelling]. 

Similar to the set of questions about the built environment, these questions have two parts: first, the respndents were asked to assess their experience of some aspect of their commute, and then  the importance they assign to this aspect of their commute. For example, variable `r1A_STRESS` refers to the level of stress that respondents experience in their usual trips; its companion variable is `r1GA_IMPSTRESS`, which refers to how important stress is to their experience. As seen in @tbl-Health-Information-Descriptives, a majority of respondents rate the stress in their commutes as medium or lower. This aspect of the experience is rated as very important or important by a majority of respondents. Other variables in this table contain information about the physical effort involved in their usual trips (`r1B_EFFORT` and `r1GB_IMPEFFORT`); proximity to other travelers in their usual trips (`r1C_PROX` and `r1GC_IMPPROX`); the environmental pollution they are exposed to in their usual trips (`r1D_CONTAM` and `r1GD_IMPCONTAM`); how safe they feel in their usual trips (`r1E_SAFETY` and `r1GE_IMPSAFETY`); and finally how comfortable their trips are (`r1F_COMFORT` and `r1GF_IMPCOMFORT`). 
```{r summary-health-information}
#| tbl-cap: Variables regarding health information of respondents
#| tbl-cap-location: top
#| label: tbl-Health-Information-Descriptives
#| echo: false
#
#skim(Santiago_H)
skim_SH <- Santiago_H |>
  select(-ID) |>
  skim()
skim_SH$factor.top_counts[1] <- skim_SH$factor.top_counts[1] |>
  str_replace_all(c(HIG = "HIGH",
                    VHI = "VERY HIGH", 
                    MED = "MEDIUM",
                    LOW = "LOW",
                    VLO = "VERY LOW"))

skim_SH$factor.top_counts[2] <- skim_SH$factor.top_counts[2] |>
  str_replace_all(c(NEU = "NEUTRAL", 
                    POS = "POSITIVE",
                    VPO = "VPOSITVE",
                    NEG = "NEGATIVE"))

skim_SH$factor.top_counts[3] <- skim_SH$factor.top_counts[3] |>
  str_replace_all(c(POO = "POOR", 
                    GOO = "GOOD",
                    FAI = "FAIR",
                    EXC = "EXCELLENT"))

skim_SH$factor.top_counts[4] <- skim_SH$factor.top_counts[4] |>
  str_replace_all(c(VDI = "VDISSATISFIED", 
                    DIS = "DISSATISFIED",
                    UNS = "UNSURE",
                    SAT = "SATISFIED"))

skim_SH$factor.top_counts[5] <- skim_SH$factor.top_counts[5] |>
  str_replace_all(c(VDI = "VDISSATISFIED", 
                    DIS = "DISSATISFIED",
                    UNS = "UNSURE",
                    SAT = "SATISFIED"))

skim_SH$factor.top_counts[6] <- skim_SH$factor.top_counts[6] |>
  str_replace_all(c(VDI = "VDISSATISFIED", 
                    DIS = "DISSATISFIED",
                    UNS = "UNSURE",
                    SAT = "SATISFIED"))

skim_SH$factor.top_counts[7] <- skim_SH$factor.top_counts[7] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH$factor.top_counts[8] <- skim_SH$factor.top_counts[8] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH$factor.top_counts[9] <- skim_SH$factor.top_counts[9] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH$factor.top_counts[10] <- skim_SH$factor.top_counts[10] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH$factor.top_counts[11] <- skim_SH$factor.top_counts[11] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH$factor.top_counts[12] <- skim_SH$factor.top_counts[12] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SH
```

**NOTE FROM ANTONIO:** SEVERAL TABLES NEED MY ATTENTION BEFORE THEY ARE READY TO BE SEEN BY BEA

<!--

@Mella2021drivers

First part of the data set of feelings and emotions (see @tbl-Feelings-Emotions-Descriptives) reveals factorial variables that people consider them pertained to a specific mode of travel. In terms of freedom, health and social interactions most of people mention them related to walking as a transportation mode. Many people consider using bus connected to unsafety, poverty, unpunctuality, congestion and uncomfortale conditions variables.In terms of using car as the mode of travel, most of respondents identified it connected to functionality, safety, luxury and status. A large group of people declare that bike riding would be related to enjoyment, low-cost and environment care. The only efficiency variable is being connected to using metro as the mode of transportation in respondents' opinions. In the second part of this table people evaluate the level of enjoy when they are traveling to their daily activities and most of them use average level for this variable and revealed the quality of their trips are very low which is not satisfying. Totally in this table we have half of respondents' answer and the other half is missing values in all variables.
feelings and emotions information can be used to mapping and understanding travel behavior and would lead to a more sustainable transportation network. Additionally, this information can make policy makers aware to make targeted choices about where to make physical amelioration as well as on which characteristics to attempt to influence perceptions of alternative modes[@anable2005all]. Similarly, by trying to identify a correlation between feelings, emotions and modes of transportation, planners would be able to persuade people to use more active modes of travel or public transport will be resulted in making our transportation system more sustainable. 

```{r summary-feelings-and-emotions}
#| tbl-cap: Variables regarding feelings and emotions of respondents
#| tbl-cap-location: top
#| label: tbl-Feelings-Emotions-Descriptives
#| echo: false
#
#skim(Santiago_FE)
skim_SFE <- Santiago_FE |>
  select(-ID) |>
  skim()
skim_SFE$factor.top_counts[1] <- skim_SFE$factor.top_counts[1] |>
  str_replace_all(c(WAL = "WALK", 
                    BIC = "BICYCLE",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[2] <- skim_SFE$factor.top_counts[2] |>
  str_replace_all(c(MOT = "MOTO",
                    WAL = "WALK", 
                    BIC = "BICYCLE"))

skim_SFE$factor.top_counts[3] <- skim_SFE$factor.top_counts[3] |>
  str_replace_all(c(MET = "MOTO",
                    BIC = "BICYCLE",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[4] <- skim_SFE$factor.top_counts[4] |>
  str_replace_all(c(BIC = "BICYCLE",
                    WAL = "WALK",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[5] <- skim_SFE$factor.top_counts[5] |>
  str_replace_all(c(BIC = "BICYCLE",
                    WAL = "WALK",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[6] <- skim_SFE$factor.top_counts[6] |>
  str_replace_all(c(WAL = "WALK",
                    BIC = "BICYCLE",
                    COL = "COLECTIVO"))

skim_SFE$factor.top_counts[7] <- skim_SFE$factor.top_counts[7] |>
  str_replace_all(c(MET = "METRO",
                    TAX = "TAXI",
                    WAL = "WALK"))

skim_SFE$factor.top_counts[8] <- skim_SFE$factor.top_counts[8] |>
  str_replace_all(c(MET = "METRO",
                    WAL = "WALK"))

skim_SFE$factor.top_counts[9] <- skim_SFE$factor.top_counts[9] |>
  str_replace_all(c(MET = "METRO",
                    WAL = "WALK"))

skim_SFE$factor.top_counts[10] <- skim_SFE$factor.top_counts[10] |>
  str_replace_all(c(MET = "METRO",
                    TAX = "TAX"))

skim_SFE$factor.top_counts[11] <- skim_SFE$factor.top_counts[11] |>
  str_replace_all(c(MET = "METRO",
                    BIC = "BICYCLE",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[12] <- skim_SFE$factor.top_counts[12] |>
  str_replace_all(c(TAX = "TAXI",
                    BIC = "BICYCLE",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[13] <- skim_SFE$factor.top_counts[13] |>
  str_replace_all(c(BIC = "BICYCLE",
                    WALK ="WALK",
                    MET = "METRO"))

skim_SFE$factor.top_counts[14] <- skim_SFE$factor.top_counts[14] |>
  str_replace_all(c(WALK ="WALK",
                    BIC = "BICYCLE",
                    TAX = "TAXI"))

skim_SFE$factor.top_counts[15] <- skim_SFE$factor.top_counts[15] |>
  str_replace_all(c(WALK ="WALK",
                    MET = "METRO",
                    BIC = "BICYCLE"))

skim_SFE$factor.top_counts[16] <- skim_SFE$factor.top_counts[16] |>
  str_replace_all(c(MET ="METRO",
                    COL = "COLECTIVO",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[17] <- skim_SFE$factor.top_counts[17] |>
  str_replace_all(c(BIC ="BICYCLE",
                    WAL = "WALK",
                    MOT = "MOTO"))

skim_SFE$factor.top_counts[18] <- skim_SFE$factor.top_counts[18] |>
  str_replace_all(c(TAX ="TAXI",
                    BIC = "BICYCLE",
                    WAL = "WALK"))

skim_SFE$factor.top_counts[19] <- skim_SFE$factor.top_counts[19] |>
  str_replace_all(c(MED = "MEDIUM", 
                    HIG = "HIGH",
                    VHI = "VERY High",
                    LOW = "LOW"))

skim_SFE$factor.top_counts[20] <- skim_SFE$factor.top_counts[20] |>
  str_replace_all(c(VLO = "VERY LOW", 
                    LOW = "LOW",
                    MED = "MEDIUM",
                    HIG = "HIGH"))

skim_SFE$factor.top_counts[21] <- skim_SFE$factor.top_counts[21] |>
  str_replace_all(c(ALL = "ALL OF THEM", 
                    TRA = "TRAVEL TIME",
                    CRO = "CROWDNESS OF PASSENGERS",
                    COM = "COMFORT ABSENCE"))

skim_SFE$factor.top_counts[22] <- skim_SFE$factor.top_counts[22] |>
  str_replace_all(c(ALL = "ALL OF THEM", 
                    RED = "REDUCTION OF TIME TRAVEL",
                    LES = "LESS CROWDNESS OF PASSENGERS",
                    BET = "BETTER QULALITY ON STREETS"))

skim_SFE
```

Table of decision-making and planning of respondents (see @tbl-Reason-Planning-Decision-Descriptives) includes different factorial variables using Likert scale and has two parts of an assessment of a variable and its importance from respondents' viewpoints. Most of people assign level of good when they assess their access to employment opportunities through public transport and presume a good level about their access to public transport which allows them to access the employment they need. While people sometimes visit family and friends, do recreational, cultural and sport activities,  and they assign a  moderate importance to them, most of them often go for grocery/food shopping and social activities and consider it moderately important. In terms of options most of people assume it is very important to have several options in using different modes of transport and they assign very high when they consider quality of life depends on the access they currently have to public transport. Also most people highly think their quality of life would increase if they have better access to public transport.About the affordability and unaffordability of a mode of travel, a large group of people assign car and taxi to these variables, respectively. For different aspects of the public transport system most of people mention that it has very importance for them to improve access to offices and commercial areas, disponibility of different transport modes, comfort for the use of public transport and the incorporation of other modes to the fare system. Almost all the variables are fairly complete except being economic and uneconomic variables which nearly half of respondents reacted to them.

The most important role of the transportation network and public modes of travel is to provide people with access to different destination in order to travel for business, reuniting with other people, doing grocery shopping and so on. This data set gives us a wide range of variables helpful for identifying people's travel pattern and what level of importance do people assign.So it would be useful for transport-related experts to have an insight about people movement to provide network system with a an appropriate level of performance to increase the efficiency and promote urban transportation system.
```{r summary-reason-and-planning-decision}
#| tbl-cap: Variables regarding decision-making and planning of respondents
#| tbl-cap-location: top
#| label: tbl-Reason-Planning-Decision-Descriptives
#| echo: false
#
#skim(Santiago_RPD)
skim_SRPD <- Santiago_RPD |>
  select(-ID) |>
  skim()

skim_SRPD$factor.top_counts[1] <- skim_SRPD$factor.top_counts[1] |>
  str_replace_all(c(GOO = "GOOD", 
                    FAI = "FAIR",
                    VER = "VERY GOOD",
                    POO = "POOR"))

skim_SRPD$factor.top_counts[2] <- skim_SRPD$factor.top_counts[2] |>
  str_replace_all(c(GOO = "GOOD",
                    EXC = "POOR",
                    VER = "VERY GOOD", 
                    FAI = "FAIR"))

skim_SRPD$factor.top_counts[3] <- skim_SRPD$factor.top_counts[3] |>
  str_replace_all(c(SOM = "SOMWTIMEs",
                    OFT = "OFTEN",
                    ALW = "ALWAYS", 
                    NEV = "NEVER"))
 
skim_SRPD$factor.top_counts[4] <- skim_SRPD$factor.top_counts[4] |>
  str_replace_all(c(SOM = "SOMWTIMEs",
                    OFT = "OFTEN",
                    ALW = "ALWAYS", 
                    RAR = "RARELY"))

skim_SRPD$factor.top_counts[5] <- skim_SRPD$factor.top_counts[5] |>
  str_replace_all(c(SOM = "SOMWTIMEs",
                    RAR = "RARELY",
                    NEV = "NEVER",
                    OFT = "OFTEN"))

skim_SRPD$factor.top_counts[6] <- skim_SRPD$factor.top_counts[6] |>
  str_replace_all(c(SOM = "SOMWTIMEs",
                    NEV = "NEVER",
                    RAR = "RARELY",
                    OFT = "OFTEN"))

skim_SRPD$factor.top_counts[7] <- skim_SRPD$factor.top_counts[7] |>
  str_replace_all(c(OFT = "OFTEN",
                    SOM = "SOMWTIMEs",
                    ALW = "ALWAYS",
                    NEV = "NEVER"))

skim_SRPD$factor.top_counts[8] <- skim_SRPD$factor.top_counts[8] |>
  str_replace_all(c(OFT = "OFTEN",
                    SOM = "SOMWTIMEs",
                    ALW = "ALWAYS",
                    NEV = "NEVER"))

skim_SRPD$factor.top_counts[9] <- skim_SRPD$factor.top_counts[9] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT",
                    VER = "VERY IMPORTANT"))

skim_SRPD$factor.top_counts[10] <- skim_SRPD$factor.top_counts[10] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SRPD$factor.top_counts[11] <- skim_SRPD$factor.top_counts[11] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SRPD$factor.top_counts[12] <- skim_SRPD$factor.top_counts[12] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[13] <- skim_SRPD$factor.top_counts[13] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT",
                    VER = "VERY IMPORTANT"))

skim_SRPD$factor.top_counts[14] <- skim_SRPD$factor.top_counts[14] |>
  str_replace_all(c(MOD = "MODERATELY IMPORTANT",
                    IMP = "IMPORTANT",
                    NOT = "NOT IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[15] <- skim_SRPD$factor.top_counts[15] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "NOT IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[16] <- skim_SRPD$factor.top_counts[16] |>
  str_replace_all(c(HIG = "HIGH",
                    VHI = "VERY HIGH",
                    MED = "MEDIUM",
                    LOW = "BELOW AVERAGE"))

skim_SRPD$factor.top_counts[17] <- skim_SRPD$factor.top_counts[17] |>
  str_replace_all(c(HIG = "HIGH",
                    VHI = "VERY HIGH",
                    MED = "MEDIUM",
                    LOW = "BELOW AVERAGE"))

skim_SRPD$factor.top_counts[18] <- skim_SRPD$factor.top_counts[18] |>
  str_replace_all(c(MET = "METRO"))

skim_SRPD$factor.top_counts[19] <- skim_SRPD$factor.top_counts[19] |>
  str_replace_all(c(TAX = "TAXI",
                    MOT = "MOTO",
                    COL = "COLECTIVO"))

skim_SRPD$factor.top_counts[20] <- skim_SRPD$factor.top_counts[20] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "NOT IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[21] <- skim_SRPD$factor.top_counts[21] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "NOT IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[22] <- skim_SRPD$factor.top_counts[22] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "NOT IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD$factor.top_counts[23] <- skim_SRPD$factor.top_counts[23] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "NOT IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SRPD
```

It can be seen there are different variables in nature and sustainability table (see @tbl-Nature-Nustainability-Descriptives) more through using Likert scale to organize respondents' answer  to variables.We can see lots of missing values for changing the mode because most of people do not tend to change their main mode  of travel when it comes to a climatic event like heavy rain or flood and for whom wants to change, using car has the most priority. Most of people have poor level of access to the currently available sustainable modes of transport (eg hybrid buses, electric cars, public bicycles) and they assign high level of importance to that. Again we have so many missing value in payment variable because almost half of people would be willing to spend more on transportation to gain access to more sustainable modes and they indicate 5-15% of their payments they tend to spend in this regard.In terms of level of importance of improving different aspects in public transport routes most respondents assume high level of importance to presence of trees,  access to parks, access to sustainable transport modes and broaden supply of sustainable transport modes. 
Nature and sustainablity information will helps planner to provide people with a safe, convenient and environment-friendly transportation system. This will also boost the quality of life, people's mobility and goods and enhance economic growth through efficient transportation services. Moreover policy makers would put in place environmental laws and excessive charges for personal car use and ownership [@alyavina2020mobility].

```{r summary-nature-and-sustainability}
#| tbl-cap: Variables regarding perspectives about nature and sustainability of respondents
#| tbl-cap-location: top
#| label: tbl-Nature-Nustainability-Descriptives
#| echo: false
#
#skim(Santiago_NS)
skim_SNS <- Santiago_NS |>
  select(-ID) |>
  skim()

skim_SNS$factor.top_counts[2] <- skim_SNS$factor.top_counts[2] |>
  str_replace_all(c(MET = "METRO", 
                    TAX = "TAXI",
                    COL = "COLECTIVO"))

skim_SNS$factor.top_counts[3] <- skim_SNS$factor.top_counts[3] |>
  str_replace_all(c(POO = "POOR",
                    FAI = "FAIR",
                    GOO = "GOOD", 
                    VER = "VERY GOOD"))

skim_SNS$factor.top_counts[4] <- skim_SNS$factor.top_counts[4] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SNS$factor.top_counts[6] <- skim_SNS$factor.top_counts[6] |>
  str_replace_all(c("30" = "30% or more",
                    "15" = "15-30%",
                    "5-1" = "5-15%")) 
                  
skim_SNS$factor.top_counts[7] <- skim_SNS$factor.top_counts[7] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    NOT = "NOT IMPORTANT"))

skim_SNS$factor.top_counts[8] <- skim_SNS$factor.top_counts[8] |>
  str_replace_all(c(VER = "VERY IMPORTANT",
                    IMP = "IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SNS$factor.top_counts[9] <- skim_SNS$factor.top_counts[9] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))

skim_SNS$factor.top_counts[10] <- skim_SNS$factor.top_counts[10] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))
                    
                    
skim_SNS
```
-->

## Experimental Design, Materials and Methods

The study is based on a paper-based survey conducted face-to-face in Santiago in 2016. The survey collected information on a wide range of travel-related issues (socio-demographics, health-related, perceptions and travel behavior, travel choices and planning, social interaction factors, built environment, among others). The data collection considered a quota-sampling method based on the information from Pre-Census of 2012, and in total, 451 persons validly completed the survey. This paper considers the first part of the survey, with information about the basic socio-economic data, travel choices, activities and commuting information, and the question related to the levels of stress experienced in while traveling.

<!--chapter:end:01-chap1.Rmd-->

---
output:
  pdf_document: default
  html_document: default
---

<!--
# Is your commute like a bad boss? Learned helplessness and normalization of stress in the commute experience in Santiago
-->
```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
#Load packages:
library(ggthemes)
library(kableExtra)
library(knitr)
library(readxl)
library(tidyverse)
library(bSantiago) # A data package with the results of a travel and well-being survey conducted in Santiago in 2016
library(dplyr) # A Grammar of Data Manipulation
library(skimr) # Compact and Flexible Summaries of Data
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(ggplot2)
library(reshape2)
library(plyr)
library(ggspatial)
library(sf)
library(shadowtext)
library(ggmosaic)
# install.packages("devtools")
devtools::install_github("haleyjeppson/ggmosaic")
```

```{r}
comunas_1 <- read_sf('C:/Users/nalaeen/Desktop/Master-Research/bSantiago/data-raw/comunas_1')
```
# Introduction

Travel essentially could have an advantage which can be different from reaching a destination or for achieving a goal. This phenomenon would be interpreted as an indirect travel in which travel has an independent entity from the activity. Moreover, people may undertake excess travel even combined with mandatory trips and do not try to make it reasonable in accordance with the target of travel. Other psychological aspects like the joy and relaxation may have an important role from their viewpoint [@mokhtarian2001derived].

According to people’s target of travel, people will assign importance on two types of psychological motives namely as instrumental (environment, cost, health and fitness, convenience, predictability, flexibility) and affective (Relaxation, no stress, excitement, control, freedom). For instance, if they are traveling for a leisure time, they expect more freedom, convenience and low level of stress. For work trips, while active modes’ users were satisfied with these motives almost equally, users of motorised modes are satisfied with instrumental factors with higher rate than affective factors. For these kinds of users, convenience and flexibility may outweigh affective factors such as stress as previous study has shown [@anable2005all]. This can be a kind of normalization because people try to adjust what they are struggling with and focus on other related dimensions to the problem. But there is a question: Should they do this in only life-threading conditions, or it works also in non-life-threading conditions? Previous work has shown that if people use normalization in non-life-threading condition it may impair the level of their quality of life.

As earlier studies have shown commuting experience may be accompanied by work-related stressors [@koslowsky1997commuting]. Interestingly, People are different in their sensitivity and vulnerability to stressful events [@shamoa2010aggression]. At high risk of experiencing stress, people are in danger of feeling psychological problems and sleep disturbance. Individuals who perceived themselves as physically weak had higher rates of mental health problems. This would be an apparent sign of the importance that people would assign to conditions in which they feel psychological problems leading to lose their self-esteem, face with self doubt and feel stress. The unremitting stress could trigger psychological issues of anxiety, fear, panic attacks, posttraumatic stress symptoms, psychological distress, stigma, avoidance of contact, depressive tendencies, sleep disturbances, helplessness and interpersonal social isolation depending on their profession. In some cases, the correlation between psychological resilience level and the perceived stress level is turned out to be reverse in a way that they may feel more resilient to the psychological effects like stress and they perceive less stress level. It can be seen in some positions Fear of labeling, stigmatization and discrimination potentially impede workers intent to seek counselling and psychotherapeutic interventions. Despite the common mental health problems and psychosocial issues among workers in such settings, most of them do not often seek or receive a systematic mental health care. 

According to a theory based on Lazarus and his colleagues, stress is perceived as a relationship between an individual and the environment relevant to its wellbeing in which they have positive or negative evaluation. Having this feeling, cognitive appraisal has two parts: first a person can decide whether the encounter is irrelevant, benign-positive or stressful. In second part the person will measure coping resources and alternatives, try to address his/her question about what they should do in this situation. These parts work interdependently in a way that if coping strategies are considered as enough resources, the threat of the event will diminish and vice versa. 

Emotions are of importance as people interpret their perceptions toward what is going on in an environment. As a diagnostic tool, emotions are vital because their intensity and quality determine how people organize the feeling of importance to them and this has a direct relation with their evaluation of a happening. 

In general, coping strategies has two majors namely as cognitive and behavioral besides a broad range of functions to handle a problematic situation between a person and an environment. First major is centered around the control of agitating emotions which is more practical when the encounter is unchangeable from a person’s perception (emotion-based coping). Some kinds of emotion-based coping strategies are identified as diminishing threat, seeking emotional or social support, wishful thinking and self-blame. The second one seeks to improve the annoying situation by doing an act (problem-based coping). In contrast, problem-based coping strategies are more popular when the encounter is changeable.  In the following paragraphs eight scales of these strategies will be identified by their own characteristics.

Wishful thinking is a way that focuses on how a person is dealing with the problem by thinking about changing the situation or tackle with that like maintaining a normal life, thinking about solutions, maintaining situational control and information seeking. Distancing as the second one refers to trying to forget the whole matter and waiting for what will happen next. Emphasizing the positive would be the third one focusing on considering the bright side of things like using a positive altitude while addressing the issue. Items four to six contain self-blame (criticizing myself), tension-reduction (trying to make a better feeling by doing favorite behaviors and self-isolation (avoidance of contact with others as the negative copying style and related to high levels of stress. All the others are associated with lower levels). Seventh item is a mix of both problem and emotion-based coping entitled “Seeking social support” covers sympathetic feelings and talking to someone else in order to making the situation easier to understand. This is the one used typically by people more than choosing one form or the other one [@folkman1985if]. Apart from classification, another kind of emotion-based strategies implies adaptive coping strategies including religion and social support [@babore2020psychological]. 

The Last item as a salient statement of problem-based coping strategies would be “adhere to a plan of actions” [@folkman1985if]. Depending on the strategies which people take into account for coping with mobility stress, vulnerability can be indicated. Mobility stress could be a situation in which people are unable to use comfortable means of mobility. Vulnerability means a living situation which is detrimental to psychological condition of people. Regarding this people may adopt strategies such as: travel by non-motorised means, readiness to high fare, cancellation of trips, travel in company of relatives, reduced number of trips, purchase of private automobile [@odufuwa2008gender].  Another coping strategy would be time allowing the individual to control stressors related to time components. Similarly, increasing control or predictability may be influential in this regard. If the commuter can be informed of special situations that exist on the way to work, such as traffic accidents or a closed road, both perception of control and predictability could be enhanced. Interestingly, with the aid of radio station in the morning (in many communities today special radio frequences in the morning and evening report on traffic patterns and even make recommendations on alternative routes) the expected negative effects of commuting can be mitigated. For example, flexitime, telecommuting, and subsidised carpooling are techniques for making the commute easier [@koslowsky1997commuting]. 

The objective of this research is to investigate whether normalization/learned helplessness could be coping strategies, and if so for whom. Data for the research are drawn from a survey conducted in Santiago, Chile, based on a quota-sampling method based on the information from Pre-Census of 2012, and in total, 451 persons validly completed the survey.

The rest of the paper is structured as follows…

# Background

## Stress and transportation

Generally, stress include a wide variety of factors and there is no single identifier broadly accepted. One of these would be stress caused by commuting. People with any target of travel may face stress while commuting between origins and destinations. Although a broad range of source of stress are identified as commuting stress, they can be primarily classified as two categories of objective stressors and subjective moderators.

 The first group indicates objectives as impedance like commuting time, distance or speed as a combination of time and distance, and commuting condition such as traffic congestion. Previous studies reveal that length of travel and congestion are strongly related to the increase of stress. People feel less stress when they experience a movement with less possible amount of time and distance. (The concept of commute impedance was specified as a behavioral control on movement or reaching a target. In other words, it contains anything that impose frustration on the goal to be reached at a certain time at a particular place such as distance, slow speed or traffic congestion. Speed would not have direct effect as an impedance, but some other extraneous factors may split from that. To make a clarification on that we can assume two drivers first driving at 35 mile/hour and second driving at 50 mile/hour. The former one may be driving on local roads, at maximum possible speed and without experiencing any barrier while the former one may be driving on a highway, face congestion (on superhighways like German autobahn) and find negative stimuli on the road [@koslowsky1997commuting].
 
Second cluster contains subjective variables like perception of control over the commute (commuting mode), predictability of commuting conditions and personal characteristics namely as gender or family specifications.  
While there is no evidence indicating effect of subjective factors, some studies have considered active modes like driving a car or riding a bike which provide more control, are more associated with less stress that passive modes like public transport. On the contrary, another study hypothesize that car users are more at the exposure of stress than bus or train users. Given these, there might be a possibility of other factors like noise or congestion to have effect on perceived stress by different modes of transportation.

As a kind of moderator, predictability of commuting has a significant role in perceived stress on commuting. Commuters capable of predicting the length of travel may face less commuting stress than their counter parts who are uncertain about this. Also, the relation of variability of commute and higher level of commuting stress is evident. 

Also, there are other factors can be considered to show the effect of stress on commuting such as assessment of commuting time, alcohol consumption, leisure-time activities, and job attributes [@gottholmseder2009stress]

## Coping with stress: Normalization and learned helplessness

The concept of normalization was mostly developed in nursing research on families with chronically ill children. Normalization consists of reducing the feeling of being different and achieving as normal a life as possible,  acknowledging the existence of the impairment (which distinguishes it from denial), using a normalcy lens to define the situation, constructing a story of life as normal and going on to enact it, and carrying on with life activities as if the illness can be ignored.

People at high risk of experiencing stress like health workers are in danger of feeling psychological problems and sleep disturbance. By using some scales for instance resilience scale and perceived stress scale including Likert scale, through a survey we can figure out to what extent people can tolerate stressful conditions and what is their exact perception of this feeling. It has been stated that individuals who perceived themselves as physically weak had higher rates of mental health problems. This would be an apparent sign of the importance that people would assign to conditions in which they feel psychological problems leading to lose their self-esteem, face with self doubt and feel stress.

It has been concluded from previous studies that people regarding their profession might be at high risk of anxiety, stress, and depression. The severity is causing further mental health problems which not only effect workers’ decision-making ability but could also have long term detrimental effect on their overall well-being. The risk of a sudden change in their roles like being downgraded can lead to psychological problems, such as disappointment, helplessness, adaptation problems, and fear of change. It seems people due to helplessness may use some coping strategies such as leaving the stressful situation and asking social support. When individuals feel they are in an uncontrolled situation they are more likely to revert to avoidance-focused coping strategy. Those victims who could not deal with these toxic conditions nor leave may feel learned helplessness and chronic health issues in long term harm [@webster2016fight]. Normalize stress in some situations due to individuals’ specific conditions may prevent them to deal with the main stressors. Similarly, people who are under high stress environment – emotional and behavioral responses are naturally adaptive in the face of extreme (unpredictable and uncertain) stress, and thus counselling and psychotherapy based on the stress-adaptation model might act as early and prompt intervention [@rana2020mental]. This feeling of helplessness might have origins in uncontrollable factors and people assume their behaviour and outcomes are independent and this learning produces the motivational, cognitive and emotional effect of uncontrollability. Learned helplessness will be shown as a defective performance where reduced incentive motivation is undermined [@maier1976learned].

When individuals are in a situation in which they suffer from a (physical) problem, depending on to what extent they have internalized the norms and principles, they might be distressed by unexpected and uncontrollable barriers to its fulfillment. In such conditions, normalization seems to play a significant role by trying to achieve higher levels of quality of life and better adjustment. But in fact, normalization is weakly related to quality of life. Normalization in long-term impairs quality of life mainly due to its social consequences. On the one hand, normalization in other non-life-threatening chronic or long-term conditions that are known to impair quality of life of young adults as they struggle to attain the developmental milestones reached by their own peers. On the other hand, Normalization includes a focus on other goals in life, instead of or in addition to those that are currently blocked. For example, by adjusting another related goal following failed treatment we can mitigate its effect on. Nevertheless, people can choose to (re)engage in other goals at the same time. Engaging in alternative meaningful goals has been found to be adaptive for people undergoing serious physical problems, with its effects seen mainly in terms of greater well-being. Normalization, in the sense of maintaining life routines and balancing between the physical and other life domains, allows for such goal re-engagement.

## Stress and coping strategies in transportation

The inevitable role of human being in the general chain of traffic issues is taken into consideration as it leads to stress, accident and aberrant behaviours. [@westerman2000individual] have revealed that there would be two predominant classification of Driver Behaviour Inventory (DBI) assessing driver stress and Driver Behaviour Questionnaire (DBQ) investigating aberrant driver behaviour. 

Unsatisfying traffic conditions may aggravate the situation and increase frustration on the road leading to high level of stress. Some people would underestimate the risk of confrontation with an accident and tend to overrate their capabilities while driving. While this high level of confidence in vehicle control decreases driver stress level, it would lead people to adopt dangerous attitudes and driving maneuvers requiring specific skills. (beneficial level of stress because it can mediate risk of stress. Stress is generally bad when it is excessive.)

The Driver Behaviour Inventory as a measure of driver stress indicates driver stress and performance as a function of evaluating traffic demands, appraising personal competence and selecting coping strategies. According to recent studies of DBI, three main aspects of driver stress vulnerability has been emerged namely as: 

•	Aggression (i.e., anger, impatience and risk taking which can be addressed through confrontive coping strategies)

•	Dislike of driving (i.e., anxiety, self-blame and lack of confidence tackled by self-criticism coping strategies or negative emotions that diverts attention from the driving task)

•	Alertness (i.e., awareness of risk and active search for road hazards issued by task-focus (problem solving) coping strategies.

*According to a study conducted by [@kontogiannis2006patterns], the direct relationship between aggression and self-criticism has been found and reveals aggressive drivers could be aware of their behaviour but they cannot take preventive actions. These findings would be beneficial for traffic safety organization looking for reducing aggression on the roads.

According to DBI, there would be two additional categories of driver stress as irritation when overtaken and frustration when overtaking, has been suggested. It seems that these overtaking factors in comparison to the former factors are more predominant on aggression. Following this, two new “situation-specific” factors in a five-factor solution were identified. “Situation-specific” tension associated with aggression and included two overtaking factors while “situation-specific” concentration correlated with alertness but consisted of only two items. Furthermore, confidence factor was recognized based on an extensive questionnaire including a new confidence factor related to control perception. An interesting relation exits between confidence and violation (aberrant behaviour: leading to accident and caused by stress) as people with high record of violations felt more confident about vehicle control. (Which is maybe related to the relation among, underrate risk of accident, overrate driving skills, high level of confidence, high possibility of undertaking driving maneuvers). [Finally, new factors of “fatigue” and “thrill seeking” have been revealing followed by an administration of a questionnaire. Studies within the context of real-world behaviour, have revealed high aggression related to faster driving and more risky overtaking. While dislike scale was associated with weaker vehicle control. An interesting study has found men reported higher aggression and comparatively lower overtaking tension. The relations between age and driver stress have indicated older drivers due to declining trend of cognitive capacities experienced high stress and were less alert  [@westerman2000individual].

Coping strategies: In stressful traffic situation, individuals should select appropriate actions to cope with stress. As mentioned before, two general categories of coping strategies have been identified: problem-based and emotion-based. 

Problem-focused followers tend to adopt tasks and rational strategies such as: 

o	Information seeking
o	Taking precaution and making plan: Making an action to drive in a safe way such as avoid making risky overtaking and changing route when it is necessary in an appropriate time
o	Confrontive coping: Expressing feelings through risk taking like driving close to next car’s bumper, beeping, flash light beams on other drivers. 
On the contrary, emotion-based drivers appear to regulate emotions and decrease discomfort by different ways such as:
o	(Escape avoidance) Avoidance: Trying to supress negative feelings to overcome the problem
o	Wishful thinking
o	Reappraisal: Considering driving as a learning experience such as trying oneself to calm down and emphasizing the positive point
o	Self-criticism: Criticizing oneself for mistakes such as feeling failure and self-doubt
The adherence of these two broad coping strategies has been recognized diverse and related to individuals’ preferences. 
Furthermore, studies about chronic stress (DBI scale) and coping behaviours has led to a consistent relationship between them. First dimension implies that drivers revealing anger, frustration and impatience (like aggression scale) were more likely to tackle stress through confrontive coping. Second would be drivers feeling anxiety and fear (like dislike of driving scale) were more interested in opting negative emotional coping styles deviating attention from driving action. Alertness as last dimension and an attribute of active hazard surveillance seems to be addressed by a problem-solving coping strategy. Alertness anticipated speed of discrimination of roadside pedestrians.
To improve the distinction of problem-focused versus emotion-focused coping skills, it is vital to define specific road scenarios and observe how drivers respond to them rather than considering general form of driving which is not enlightening (like slow-driving: someone is driving in front of you making overtaking hard, and discourtesy scenarios: someone is driving behind you near to your rear bumper beeping because . In addition, diversity and socio-cultural backgrounds around the world would affect factor structure of driver stress and its correlation to coping styles.
Previous researches have shown that professional drivers chose a group of both problem-based and emotion-based strategies. This selection may change regarding the level of frustration and anger felt in various road scenarios [@kontogiannis2006patterns].

# Materials and methods

<!-- Source of questionnaire is needed to be explained  -->
The presented data in this article were gathered purposefully through a questionnaire to capture various dimensions of urban living and mobility in a major city in the Global South. The questionnaire covers different topics such as personal characteristics, work-related travel information, and perceptions of the built environment. Other themes include social interactions while using different modes of transportation, mode switching, and the use of information technologies. The questionnaire also addresses several self-reported health outcomes, feelings and emotions related to commuting experiences, attitudes toward transportation system satisfaction, and perceptions of nature and sustainability.

This study focuses on both demographic and health-related information. To assess participants' views on stress during their commute, multiple questions were asked, including variable r1A_STRESS as one of our dependent variables, which reflects the level of stress experienced during regular trips using a 5-point Likert scale (very low, low, moderate, high, and very high). Its corresponding variable, r1GA_IMPSTRESS as the other one of our dependent variables, evaluates the importance of stress during commuting and is also assessed using a 5-point Likert scale (not important, slightly important, moderately important, important, and very important).

Furthermore, health information can be useful in understanding the effects of transportation policy decisions on public health. Identifying the specific factors that contribute to stressful modes of transportation can help transportation and public health experts make commuting a safer, more enjoyable, and less stressful experience.

 

## Data preparation

Collect the data on the two ordinal variables of interest and check for any missing data or outliers. If necessary, clean the data and recode it into ordinal categories.

-The case study for this research covered the Santiago Metropolitan as Chile’s capital and largest city containing various communes as shown in the figure. 1. The dataset used in current study obtained from a paper-based survey conducted face-to-face in Santiago in 2016. The data set contains essential socio-economic and demographic information about the respondents, as well as their built environment and behaviors commuting to work. In addition, the survey (conducted between DATE-DATE, 2016) includes information about the respondents’ feelings and emotions in relation to their commuting experience, the social experience of a variety of transportation modes, various self-assessed health questions, patterns of use of information and telecommunication technologies, and questions about sustainability and the environment. The data collection conducted a quota-sampling method based on the information from Pre-Census of 2012, and in total, 451 participants took part in the survey. This study considers the first two parts of the survey including individual characteristic and health information to figure out the relationship between personal attributes and stress related variables. 

```{r}

# ggplot() +
#   annotation_map_tile() +
#   geom_sf(data = comunas_1,
#           linewidth = 1,
#           fill = NA) +
#   geom_sf(data = comunas_1, fill = "gray", color = "black", size = 0.5, alpha = 0.5) +
#   geom_sf_label(data = comunas_1, aes(COORD_X, COORD_Y, label = paste(COMUNA,sep = ", ")),
#                 size = 2, fontface = "plain") +
#   coord_sf(crs = st_crs(comunas_1)) +
#   theme_void() 
# # Adding points of comunas
# 
# tm_shape(comunas_1) +
#   tm_polygons() +
#   tm_shape(comunas_1)+
#   tm_symbols(size = 0.5)
```

```{r}
comunas_1 <- read_sf('C:/Users/nalaeen/Desktop/Master-Research/bSantiago/data-raw/comunas_1')

San_area_plot <- ggplot() +
  geom_sf(data = comunas_1, color = "darkgray",
          aes(fill = COMUNA)) +
  scale_fill_brewer(palette = "Spectral") +  # using color brewer palette
  geom_sf(data = comunas_1, 
          color = "black", fill = NA, size = 0.7) +
  annotation_map_tile() +
  geom_sf(data = comunas_1,
          linewidth = 1,
          fill = NA) +
  annotation_north_arrow(location = "tl", 
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), 
                   height = unit(0.15, "cm")) + 
  geom_shadowtext(data = comunas_1,
                  aes(x = COORD_X, y = COORD_Y, label = COMUNA),
                  size = 3.5,
                  nudge_y = 3000,
                  nudge_x = 2000) +
  theme_void() +
  theme(legend.position = "none",
        axis.title = element_blank()) 

San_area_plot

```

## Variables

As mentioned earlier in the study, previous researches have applied various variables for implementing the ordinal regression model. In this study we employ a group of dependent and independent variables extracted from individual characteristics and health information parts of the surveys. As dependent variables we chose stress indicator of level of stress experienced by respondents while commuting and another stress criterion revealing the importance level of the stress experienced when traveling by participants. For independent variables we selected respondents age and income level, occupation, and their primary mode of transportation while regular commuting. 

To initialize our bivariate ordinal model, we took advantage of previous studies about the how was the importance of different independent variables withing individual characteristics.

Although gender looked affective on some of driver stress scales like confidence and alertness, it was rather weak to indicate gender difference in terms of coping strategies. Previous studies revealed men feel more confidence and alert that their counter parts. Regarding coping strategies, women seemed to score higher in cognitive strategies despite men who tend to use behavioral strategies. [@kontogiannis2006patterns]. Although it has been found that women were more likely to feel stress and anxiety that men, some studies has shown that women and men in a rather similar conditions felt the same level of stress (Hill and Ng Boyle, 2007).

Age sounded influential on most of aspects of driver stress vulnerability. Getting older had a positive role in a way that it increased cognitive coping strategies useful in decreasing stress level and declined negative behaviors like aggression [@kontogiannis2006patterns], thus they felt less stress. Another study indicated the relation of age with stress in a completely different way as some situation-specific tensions tend to increase by getting old until some of older ones may decide to not to drive [@hill2007driver]. Age also could be negatively related to manifestation of violation and dangerous behaviors as old drivers tend to decrease the speed due to the declining cognitive capabilities [@westerman2000individual]. 

In contrast of old people, youth were more likely to reveal their driving anger through behavioral coping strategies[@hill2007driver]. 

Based on findings of previous research on determinants of individual’s vulnerability to mobility stress, level of education and family status would be influential on the experienced stress while commuting. Also, high level of income would provide people with the opportunity of purchasing a private vehicle limiting their driving stress and uncomfortable experience with public means [@odufuwa2008gender].


```{r summary of desctiptive statistics}

# Join Santiago_H and Santiago_IC by "ID"
df <- Santiago_H %>%
  left_join(Santiago_IC, by = "ID") %>%
  # Select columns from Santiago_H and Santiago_IC
  select( r0D_AGE, r0J_INCOME, r0G_EDUCATION, r0P_MODE1, r0H_OCCUPATION, r1A_STRESS, r1GA_IMPSTRESS) %>%
  drop_na(r0D_AGE,
          r0J_INCOME,
          r0G_EDUCATION,
          r0P_MODE1,
          r0H_OCCUPATION,
          r1A_STRESS, 
          r1GA_IMPSTRESS)

skim_HIC <- df |>
  select(r0D_AGE, r0J_INCOME, r0G_EDUCATION, r0P_MODE1, r0H_OCCUPATION, r1A_STRESS, r1GA_IMPSTRESS) |>
  skim()

skim_HIC$factor.top_counts[1] <- skim_HIC$factor.top_counts[1] |>
  str_replace_all(c("35t" = "35 to 54", 
                    "18t" = "18 to 24",
                    "25t" = "25 to 34",
                    "55t" = "55 to 64"))

skim_HIC$factor.top_counts[2] <- skim_HIC$factor.top_counts[2] |>
  str_replace_all(c("977" = "977 to 1,550",
                    "238" = "2,380 or more",
                    "Les" = "Less than 423",
                    "155" = "1,550 to 2,380"))

skim_HIC$factor.top_counts[6] <- skim_HIC$factor.top_counts[6] |>
  str_replace_all(c(HIG = "HIGH",
                    VHI = "VERY HIGH", 
                    MED = "MEDIUM",
                    LOW = "LOW",
                    VLO = "VERY LOW"))
skim_HIC$factor.top_counts[7] <- skim_HIC$factor.top_counts[7] |>
  str_replace_all(c(IMP = "IMPORTANT",
                    VER = "VERY IMPORTANT",
                    MOD = "MODERATELY IMPORTANT",
                    SLI = "SLIGHTLY IMPORTANT"))


skim_HIC


```


```{r Preparing dummy variables for desctiptive statistics}

df$Less18 <- ifelse (df$r0D_AGE == "Less18", 1, 0)
df$"18to24" <- ifelse (df$r0D_AGE == "18to24", 1, 0)
df$"25to34" <- ifelse (df$r0D_AGE == "25to34", 1, 0)
df$"35to54" <- ifelse (df$r0D_AGE == "35to54", 1, 0)
df$"55to64" <- ifelse (df$r0D_AGE == "55to64", 1, 0)
df$"65more" <- ifelse (df$r0D_AGE == "65more", 1, 0)



df$Less423 <- ifelse (df$r0J_INCOME == "Less423", 1, 0)
df$"423to639" <- ifelse (df$r0J_INCOME == "423to639", 1, 0)
df$"639to977" <- ifelse (df$r0J_INCOME == "639to977", 1, 0)
df$"977to1550" <- ifelse (df$r0J_INCOME == "977to1550", 1, 0)
df$"1550to2380" <- ifelse (df$r0J_INCOME == "1550to2380", 1, 0)
df$"2380more" <- ifelse (df$r0J_INCOME == "2380more", 1, 0)


df$"ELEMENTARY" <- ifelse (df$r0G_EDUCATION == "ELEMENTARY", 1, 0)
df$"SECUNDARY" <- ifelse (df$r0G_EDUCATION == "SECUNDARY", 1, 0)
df$"PROFESS_TECH" <- ifelse (df$r0G_EDUCATION == "PROFESS_TECH", 1, 0)
df$"COLLEGE" <- ifelse (df$r0G_EDUCATION == "COLLEGE", 1, 0)
df$"POSTGRAD" <- ifelse (df$r0G_EDUCATION == "POSTGRAD", 1, 0)


df$"CAR" <- ifelse (df$r0P_MODE1 == "CAR", 1, 0)
df$"TAXI" <- ifelse (df$r0P_MODE1 == "TAXI", 1, 0)
df$"COLECTIVO" <- ifelse (df$r0P_MODE1 == "COLECTIVO", 1, 0)
df$"MOTO" <- ifelse (df$r0P_MODE1 == "MOTO", 1, 0)
df$"METRO" <- ifelse (df$r0P_MODE1 == "METRO", 1, 0)
df$"BUS" <- ifelse (df$r0P_MODE1 == "BUS", 1, 0)
df$"BICYCLE" <- ifelse (df$r0P_MODE1 == "BICYCLE", 1, 0)
df$"WALK" <- ifelse (df$r0P_MODE1 == "WALK", 1, 0)


df$"FULL_TIME" <- ifelse (df$r0H_OCCUPATION == "FULL_TIME", 1, 0)
df$"PART_TIME" <- ifelse (df$r0H_OCCUPATION == "PART_TIME", 1, 0)
df$"SELF_EMPLOYED" <- ifelse (df$r0H_OCCUPATION == "SELF_EMPLOYED", 1, 0)
df$"UNEMPLOYED" <- ifelse (df$r0H_OCCUPATION == "UNEMPLOYED", 1, 0)
df$"HOME_TAKER" <- ifelse (df$r0H_OCCUPATION == "HOME_TAKER", 1, 0)
df$"STUDENT" <- ifelse (df$r0H_OCCUPATION == "STUDENT", 1, 0)
df$"STUDENT_WORK" <- ifelse (df$r0H_OCCUPATION == "STUDENT_WORK", 1, 0)
df$"RETIRED" <- ifelse (df$r0H_OCCUPATION == "RETIRED", 1, 0)
df$"OTHER" <- ifelse (df$r0H_OCCUPATION == "OTHER", 1, 0)


df$"VLOW" <- ifelse (df$r1A_STRESS == "VLOW", 1, 0)
df$"LOW" <- ifelse (df$r1A_STRESS == "LOW", 1, 0)
df$"MODERATE" <- ifelse (df$r1A_STRESS == "MODERATE", 1, 0)
df$"HIGH" <- ifelse (df$r1A_STRESS == "HIGH", 1, 0)
df$"VHIGH" <- ifelse (df$r1A_STRESS == "VHIGH", 1, 0)


df$"NOT IMPORTANT" <- ifelse (df$r1GA_IMPSTRESS == "NOT IMPORTANT", 1, 0)
df$"SLIGHTLY IMPORTANT" <- ifelse (df$r1GA_IMPSTRESS == "SLIGHTLY IMPORTANT", 1, 0)
df$"MODERATELY IMPORTANT" <- ifelse (df$r1GA_IMPSTRESS == "MODERATELY IMPORTANT", 1, 0)
df$"IMPORTANT" <- ifelse (df$r1GA_IMPSTRESS == "IMPORTANT", 1, 0)
df$"VERY IMPORTANT" <- ifelse (df$r1GA_IMPSTRESS == "VERY IMPORTANT", 1, 0)

```


```{r desctiptive statistics}

df

example_2.df <- data.frame(Variable = c("Age 1",
                                        "Age 2",
                                        "Age 3",
                                        "Age 4",
                                        "Age 5",
                                        "Age 6",
                                        "Income 1",
                                        "Income 2",
                                        "Income 3",
                                        "Income 4",
                                        "Income 5",
                                        "Income 6",
                                        "Education 1",
                                        "Education 2",
                                        "Education 3",
                                        "Education 4",
                                        "Education 5",
                                        "Primary Mode 1",
                                        "Primary Mode 2",
                                        "Primary Mode 3",
                                        "Primary Mode 4",
                                        "Primary Mode 5",
                                        "Primary Mode 6",
                                        "Primary Mode 7",
                                        "Primary Mode 8",
                                        "Occupation 1",
                                        "Occupation 2",
                                        "Occupation 3",
                                        "Occupation 4",
                                        "Occupation 5",
                                        "Occupation 6",
                                        "Occupation 7",
                                        "Occupation 8",
                                        "Occupation 9",
                                        "Experienced Stress 1",
                                        "Experienced Stress 2",
                                        "Experienced Stress 3",
                                        "Experienced Stress 4",
                                        "Experienced Stress 5",
                                        "Importance of Stress 1",
                                        "Importance of Stress 2",
                                        "Importance of Stress 3",
                                        "Importance of Stress 4",
                                        "Importance of Stress 5"
                                    
), 
Note = c("Dummy variable: 1 if Age Less18",
         "Dummy variable: 1 if Age 18-24 years",
         "Dummy variable: 1 if Age 25-34 years",
         "Dummy variable: 1 if Age 35-54 years",
         "Dummy variable: 1 if Age 55-64 years",
         "Dummy variable: 1 if Age 65 and more",
         "Dummy variable: 1 if Income Less423",
         "Dummy variable: 1 if Income 423to639",
         "Dummy variable: 1 if Income 639to977",
         "Dummy variable: 1 if Income 977to1550",
         "Dummy variable: 1 if Income 1550to2380",
         "Dummy variable: 1 if Income 2380more",
         "Dummy variable: 1 if Education ELEMENTARY",
         "Dummy variable: 1 if Education SECUNDARY",
         "Dummy variable: 1 if Education PROFESS_TECH",
         "Dummy variable: 1 if Education COLLEGE",
         "Dummy variable: 1 if Education POSTGRAD",
         "Dummy variable: 1 if Primary Mode CAR",
         "Dummy variable: 1 if Primary Mode TAXI",
         "Dummy variable: 1 if Primary Mode COLECTIVO",
         "Dummy variable: 1 if Primary Mode MOTO",
         "Dummy variable: 1 if Primary Mode METRO",
         "Dummy variable: 1 if Primary Mode BUS",
         "Dummy variable: 1 if Primary Mode BICYCLE",
         "Dummy variable: 1 if Primary Mode WALK",
         "Dummy variable: 1 if Occupation FULL_TIME",
         "Dummy variable: 1 if Occupation PART_TIME",
         "Dummy variable: 1 if Occupation SELF_EMPLOYED",
         "Dummy variable: 1 if Occupation UNEMPLOYED",
         "Dummy variable: 1 if Occupation HOME_TAKER",
         "Dummy variable: 1 if Occupation STUDENT",
         "Dummy variable: 1 if Occupation STUDENT_WORK",
         "Dummy variable: 1 if Occupation RETIRED",
         "Dummy variable: 1 if Occupation OTHER",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale",
         "Likert scale"
), 

Percentage = c(paste(round(sum(df$r0D_AGE %in% "Less18")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0D_AGE %in% "18to24")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0D_AGE %in% "25to34")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0D_AGE %in% "35to54")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0D_AGE %in% "55to64")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0D_AGE %in% "65more")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "Less423")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "423to639")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "639to977")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "977to1550")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "1550to2380")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0J_INCOME == "2380more")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0G_EDUCATION == "ELEMENTARY")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0G_EDUCATION == "SECUNDARY")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0G_EDUCATION == "PROFESS_TECH")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0G_EDUCATION == "COLLEGE")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0G_EDUCATION == "POSTGRAD")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "CAR")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "TAXI")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "COLECTIVO")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "MOTO")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "METRO")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "BUS")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "BICYCLE")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0P_MODE1 == "WALK")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "FULL_TIME")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "PART_TIME")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "SELF_EMPLOYED")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "UNEMPLOYED")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "HOME_TAKER")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "STUDENT")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "STUDENT_WORK")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "RETIRED")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r0H_OCCUPATION == "OTHER")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1A_STRESS == "VLOW")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1A_STRESS == "LOW")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1A_STRESS == "MODERATE")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1A_STRESS == "HIGH")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1A_STRESS == "VHIGH")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1GA_IMPSTRESS == "NOT IMPORTANT")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1GA_IMPSTRESS == "SLIGHTLY IMPORTANT")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1GA_IMPSTRESS == "MODERATELY IMPORTANT")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1GA_IMPSTRESS == "IMPORTANT")/nrow(df)*100, 1), "%", sep = " "),
               paste(round(sum(df$r1GA_IMPSTRESS == "VERY IMPORTANT")/nrow(df)*100, 1), "%", sep = " ")
            

              
               
))

```

```{r tabulate-descriptive-statistics, echo=FALSE}
tableout <- kable(example_2.df,
      "latex", 
      caption = "\\label{tab:descriptive-statistics}Summary statistics of the sample", 
      booktabs = T,
      digits = 1) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  pack_rows("Socio-Economic and Demographic Attributes", 1, 34) %>%
  pack_rows("Stress Characteristics", 35, 44) %>%
  row_spec(c(1, 35) - 1, extra_latex_after = "\\rowcolor{gray!20}") 

```


In order to reach the more efficient model we used recode_factor() argument which requires the dplyr R package to be loaded, and the function takes the original factor variable as the first argument, followed by the recoding rules. The recoding rules in this study are specified using a set of named character vectors to aggregate or disperse the original levels of the factor and the values represent the new levels.

```{r resetting the outcomes of each variable to see the efficient result}

df <- Santiago_H |>
  left_join(Santiago_IC,
            by = "ID") |>
  as.data.frame() |>
  drop_na(r0D_AGE,
          r0J_INCOME,
          r0G_EDUCATION,
          r0P_MODE1,
          r0H_OCCUPATION ) 
  

df <- df %>%  mutate(r0J_INCOME = recode_factor(r0J_INCOME,
                                  "Less423" = "Low Income",
                                  "423to639" = "Low Income",
                                  "639to977" = "Middle Income",
                                  "977to1550" = "Middle Income",
                                  "1550to2380" = "High Income",
                                  "2380more" = "High Income"))  %>%
  # mutate(r0D_AGE = recode_factor(r0D_AGE,
  #                                 "Less18" = "Youth and Children",
  #                                 "18to24" = "Youth and Children",
  #                                 "25to34" = "Youth and Children",
  #                                 "35to54" = "Adults",
  #                                 "55to64" = "Adults",
  #                                 "65more" = "Adults"))  %>%
    mutate(r0D_AGE = recode_factor(r0D_AGE,
                                  "Less18" = "Youth",
                                  "18to24" = "Youth",
                                  "25to34" = "Youth",
                                  "35to54" = "Adults",
                                  "55to64" = "Seniors",
                                  "65more" = "Seniors"))  %>%
  mutate(r0P_MODE1 = recode_factor(r0P_MODE1,
                                  "CAR" = "Private Transportation",
                                  "MOTO" = "Private Transportation",
                                  "TAXI" = "Public Transportation",
                                  "COLECTIVO" = "Public Transportation",
                                  "METRO" = "Public Transportation",
                                  "BUS" = "Public Transportation",
                                  "BICYCLE" = "Active Transportation",
                                  "WALK" = "Active Transportation")) %>%
  # mutate(r0P_MODE1 = recode_factor(r0P_MODE1,
  #                                 "CAR" = "Motorized Transportation",
  #                                 "MOTO" = "Motorized Transportation",
  #                                 "TAXI" = "Motorized Transportation",
  #                                 "COLECTIVO" = "Motorized Transportation",
  #                                 "METRO" = "Motorized Transportation",
  #                                 "BUS" = "Motorized Transportation",
  #                                 "BICYCLE" = "Active Transportation",
  #                                 "WALK" = "Active Transportation")) %>%
  # mutate(r0G_EDUCATION = recode_factor(r0G_EDUCATION,
  #                                 "ELEMENTARY" = "Elementary Education",
  #                                 "SECUNDARY" = "Middle Education",
  #                                 "PROFESS_TECH" = "Middle Education",
  #                                 "COLLEGE" = "Middle Education",
  #                                 "POSTGRAD" = "Higher Education"))
mutate(r0G_EDUCATION = recode_factor(r0G_EDUCATION,
                                  "ELEMENTARY" = "Elementary Education",
                                  "SECUNDARY" = "Higher Education",
                                  "COLLEGE" = "Higher Education",
                                  "PROFESS_TECH" = "Higher Education",
                                  "POSTGRAD" = "Higher Education"))  %>%
  
mutate(r0H_OCCUPATION = recode_factor(r0H_OCCUPATION,
                                  "FULL_TIME" = "Working",
                                  "PART_TIME" = "Working",
                                  "SELF_EMPLOYED" = "Working",
                                  "UNEMPLOYED" = "Not Working",
                                  "HOME_TAKER" = "Not Working",
                                  "STUDENT" = "Student",
                                  "STUDENT_WORK" = "Working",
                                  "RETIRED" = "Not Working",
                                  "OTHER" = "Non Professional"))

   # df$r0C_SEX <- relevel(df$r0C_SEX, ref = "MAN")
   df$r0P_MODE1 <- relevel(df$r0P_MODE1, ref = "Active Transportation")
   df$r0D_AGE <- relevel(df$r0D_AGE, ref = "Youth")
   df$r0J_INCOME <- relevel(df$r0J_INCOME, ref = "Low Income")
   df$r0G_EDUCATION <- relevel(df$r0G_EDUCATION, ref = "Elementary Education")
   df$r0M_DISABILITY <- relevel(df$r0M_DISABILITY, ref = "YES")
   df$r0H_OCCUPATION <- relevel(df$r0H_OCCUPATION, ref = "Not Working")
  
 
```


## Modelling approach

One of the most common analysis of ordinal data would be related to modeling the preferences or opinions in various fields. In this regard we can see the application of modelling in correlated ordinal data as multiple outcomes for a number of subjects or objects like participants responding to a questionnaire. Given these multivariate setting, models and particularly multivariate ordinal regression models sound practical to deal with the correlation among ordinal outcomes. Current study has used similar application of bivariate ordinal regression models [@kenne2016pairwise] considering the gap between expected satisfaction and experienced service. It refers to two different blocks that participants are asked to say to what extent the service has importance and as the second part they are required to reveal the actual feeling about the given service. 

As can be seen in the study conducted by [@hirk2020mvord] the class of multivariate ordinal regression models implemented in mvord and can be applied to other applications such as having multiple or repeated ordinal observations. This research takes advantage of package mvord for R Studio which provides a flexible framework for analysing correlated ordinal data using the class of multivariate ordinal regression models. This class of modeling assumes each ordinal response as a categorized version of an underlying continuous latent variable separated based on specific threshold parameters. This flexible framework contributes helps to imposing constraint on thresholds along with regression coefficients using … arguments.  

In this study we implement bivariate logit link for the class of multivariate ordinal regression models. In terms of identifiability, imposing some restriction on the parameters is needed as the absolute scale and location are not identifiable in ordinal models. Therefore, in order to make the model identifiable, we use one of the options regarding constraining parameter set by the following way:

Fixing the intercept βj0 (e.g., to zero), using flexible thresholds θj and fixing σij (e.g., to unity) ∀j ∈ Ji, ∀i ∈ {1, . . . , n}.

# Analysis and results

In order to fit our bivariate ordinal regression model, we can use mvord R package which has two different data structures of MMO and MMO2. As we are working with wide data format and the covariates stored in different columns of the data set do not change during multiple measurements, using MMO2 data structure is an appropriate way. In this data structure each subject i refers to on row of the data frame and all the covariates are stored in different columns.

At this point we used the following formula to specify the model: 

mod_bivariate <- mvord(formula = MMO2(r1A_STRESS,r1GA_IMPSTRESS) ~ 0 + r0D_AGE + r0J_INCOME + r0P_MODE1 + r0G_EDUCATION + r0H_OCCUPATION  , 
                       link = mvlogit(df = 8L)

where "mod_bivariate" is the name of the model, "mvord" is the function for fitting multivariate ordinal models, and "MMO2" is the data structure in accordance with wide format of data set. Predicators are demographic variables of the respondents namely as their age, income, education and occupation level as well as their primary mode of commute.


```{r}
library(mvord)
mod_bivariate <- mvord(formula = MMO2(r1A_STRESS, r1GA_IMPSTRESS) ~ 0 + r0D_AGE + r0J_INCOME + r0P_MODE1 + r0G_EDUCATION + r0H_OCCUPATION  ,
                       link = mvlogit(df = 8L),
                       # {mvord} does not like tbl or tbl_df objects:
                       # convert to plain data.frame
                       data = df)
summary(mod_bivariate)

```


As the second step to make sure of the model identifiability and reveal the latent correlation in the model, we used an argument to constrain the coefficients.



```{r}
formula <- MMO2(r1A_STRESS, r1GA_IMPSTRESS) ~ 0 + r0D_AGE + r0J_INCOME + r0P_MODE1 + r0G_EDUCATION + r0H_OCCUPATION  
names_constraints(formula, data = df)
coef.constraints = cbind(c(NA, 1), # Adults
                         c(1, NA), # Seniors
                         c(1, 2), # Middle Income
                         c(1, NA), # High Income
                         c(1, NA), # Private Transportation
                         c(1, NA), # Public Transportation
                         c(NA, 1), # Higher Education
                         c(1, NA), # Working
                         c(1, NA), # Student
                         c(NA, NA)) #Non Professional
# In this code 1 or 2 means yes to significant coefficient and doesn't matter if it refers to which equation. For specifying equation we have to consider which column is that.


mod_bivariate <- mvord(formula = MMO2(r1A_STRESS, r1GA_IMPSTRESS) ~ 0 + r0D_AGE + r0J_INCOME + r0P_MODE1 + r0G_EDUCATION + r0H_OCCUPATION  ,
                       link = mvlogit(df = 8L),
                       coef.constraints = coef.constraints,
                       # {mvord} does not like tbl or tbl_df objects:
                       # convert to plain data.frame
                       data = df)
summary(mod_bivariate)


```

# Discussion

The results of the model analysis can be seen in Table ... . The analysis reveals that adult individuals aged between 35 to 54 are the most affected category when it comes to rating the importance stress during commuting. There is a positive correlation indicating that adults within this age range are more likely to assign importance to stress while commuting. 

This trend holds until the age of 54, after which seniors aged 55 and above are significantly related to experience stress. The negative coefficient for this age group suggests that their stress level is decreasing compared to their younger counterparts. This finding may seem incompatible with some previous research claiming that as people get older, they tend to avoid risks and tensions, especially while driving. However, it is important to consider that older individuals may also suffer from cognitive limitations [@westerman2000individual] and have difficulties remembering their lapses, leading to potentially dangerous and stressful situations. This may put them at a higher risk of accidents or unpleasant commuting experiences with more stressful moments. The declining ability to self-monitor and remember specific errors could contribute to this phenomenon [@rabbitt1990age].

Previous research has indicated that income groups play a significant role in the relationship between travel and satisfaction. It has been suggested that commute satisfaction is associated with commute enjoyment, commute stress, social comparisons, personality, and overall well-being [@ye2019determinants]. People with higher incomes are more likely to afford better modes of travel, such as personal drivers, and therefore experience fewer stressful situations. As it can be seen in the model, the higher income category is strongly associated with a negative and relatively large coefficient, even larger than the middle-income group, indicating less probability of experiencing stress. This could be attributed to their capability or overall well-being, enabling them to choose different modes of transportation to avoid stressful commuting experiences. These individuals may utilize problem-solving or emotion-based coping strategies. As it has been indicated, the more people drive the less stress they experience as they become accustomed to it and this regularity helps them develop coping strategies and become more adept at the drive [@legrain2015stressed]. Interestingly, the middle-income group, despite having a positive coefficient, does not show a significant association with the experienced stress but it is still positive. However, it is worth noting that this group is highly and positively associated with assigning importance to the experienced stress during commute. 

Previous work has also revealed that lower-income groups (can be considered as similar to middle income?) rated instrumental factors (e.g., cost, predictability, flexibility) higher than higher income groups [@ye2019determinants]. This could explain why lower-income (middle income in our study) individuals are not strongly associated with mentioning the importance of stress. They are more likely to choose public transportation or active modes of transportation due to limited options available to them.

Moreover, it has been indicated that the impact of commuting mode on stress should be considered by transportation and planning agencies in making decisions [@wener2011comparing]. Because commuting specifies level of perceived control over this process which has a reverse relation with stress [@gottholmseder2009stress]. According to the current study, both public and private modes of transportation are positively associated with experiencing stress during commutes. Large coefficient indicates a more significant association between experiencing stress and those who choose public and private modes of transportation than active mode users. These findings are compatible with previous work indicating that commuting by car and public transit are considered to be more stressful and boring than active commuting. This has also been revealed that commuting satisfaction is associated with commute stress to a large extent and hence pedestrian, train travelers, and cyclists are significantly more satisfied with their commuting than car drivers, metro and bus users [@ye2019determinants].

Having higher level of education has been found to be associated with a positive and relatively significant coefficient leading to a more probability of assigning importance to the self-reported stress while commuting. As already has been mentioned, commute satisfaction has a strong effect on commute stress, and these are followed by short distance factor[@ye2019determinants]. According to a previous study on the effect of telework on daily travel in Sweden, people with higher education are more likely to travel longer and this finding is willing to put travelers more at the exposure of experiencing stress [@ellder2020telework].

In terms of occupation, current study has shown that experiencing stress among those who have a profession or students is significantly associated. Negative coefficient for both groups implies that their stress level has a decreasing trend compared to their counterparts. In addition, increasing trend of stress level has been indicated in some professions like construction due to different reasons like having an unsafe workplace or occupational tensions [@loosemore2004gender].

```{r}
coef.constraints <- coef.constraints |>
  t() |>
  data.frame() |>
  filter(!is.na(X1) | !is.na(X2))

# Initialize a table to present the model results with the names of the variables
mod_table <- data.frame(variable = names(mod_bivariate$beta)) |>
  # Remove the number corresponding to the coefficient constraint
  mutate(variable = str_sub(variable, end = -2)) |>
  # Unique variables
  distinct() |>
  # Now create columns with the coefficient constraints and name the variables in each of the equations
  mutate(e1 = coef.constraints$X1,
         e2 = coef.constraints$X2,
         e1_var = ifelse(!is.na(e1), paste0(variable, "1"), ""),
         e2_var = ifelse(!is.na(e1) & !is.na(e2), paste0(variable, "2"), paste0(variable, "1")),
         e2_var = ifelse(!is.na(e2), e2_var, ""))
```


<!--
```{r initialize-table-results, include=FALSE}
# Initialize a table to present the model results with the names of the variables
mod_table <- data.frame(df = names(mod_bivariate$beta)) |>
  # Remove the number corresponding to the coefficient constraint
  mutate(df = str_sub(df) |>
  # Unique variables
  distinct() |>
  # Now create columns with the coefficient constraints and name the variables in each of the equations
  mutate(e1 = coef.constraints[1,],
         e2 = coef.constraints[2,],
         e1_var = ifelse(!is.na(e1), paste0(variable, "1"), ""),
         e2_var = ifelse(!is.na(e1) & !is.na(e2), paste0(variable, "2"), paste0(variable, "1")),
         e2_var = ifelse(!is.na(e2), e2_var, ""))
```

```{r}
# Initialize a table to present the model results with the names of the variables
mod_table <- data.frame(variable = names(mod_bivariate$beta)) |>
  # Remove the number corresponding to the coefficient constraint
  mutate(variable = str_sub(variable, end = -2)) |>
  # Unique variables
  distinct() |>
  # Now create columns with the coefficient constraints and name the variables in each of the equations
  mutate(e1 = coef.constraints[1,],
         e2 = coef.constraints[2,],
         e1_var = ifelse(!is.na(e1), paste0(variable, "1"), ""),
         e2_var = ifelse(!is.na(e1) & !is.na(e2), paste0(variable, "2"), paste0(variable, "1")),
         e2_var = ifelse(!is.na(e2), e2_var, ""))

```
-->

```{r extract-estimates, include=FALSE}
# Obtain table with statistics (thresholds)
mod_theta <- mod_bivariate$theta$r1A_STRESS |>
  transmute(Estimate, `Pr(>|z|)`) |>
  rownames_to_column(var = "variable") |>
  rename(beta = Estimate,
         p = `Pr(>|z|)`)

mod_theta <- data.frame(variable = c("VLOW|LOW",
                                     "LOW|MODERATE",
                                     "MODERATE|HIGH",
                                     "HIGH|VHIGH"),
                        e1_beta = mod_theta$beta[1:5],
                        e1_p = mod_theta$p[1:5],
                        e2_beta = mod_theta$beta[6:10],
                        e2_p = mod_theta$p[6:10])

# Extract coefficient estimates and p-values
mod_beta <- mod_beta$coefficients |>
  transmute(Estimate, `Pr(>|z|)`) |>
  rownames_to_column(var = "variable") |>
  rename(beta = Estimate,
         p = `Pr(>|z|)`)
```

```{r storing thresholds (outcome variables)}
mod_theta <- data.frame(thresholds1 = c("VLOW|LOW",
                                     "LOW|MODERATE",
                                     "MODERATE|HIGH",
                                     "HIGH|VHIGH"),
                        theta1 = mod_bivariate$theta$r1A_STRESS, 
                        setheta1 = mod_bivariate$setheta$r1A_STRESS,
                        
                        thresholds2 = c("NOT IMPORTANT|SLIGHTLY IMPORTANT",
                                     "SLIGHTLY IMPORTANT|MODERATELY IMPORTANT",
                                     "MODERATELY IMPORTANT|IMPORTANT",
                                     "IMPORTANT|VERY IMPORTANT"),
                        theta2 = mod_bivariate$theta$r1GA_IMPSTRESS,
                        setheta2 = mod_bivariate$setheta$r1GA_IMPSTRESS)


```

```{r Calculating Z and P values for thresholds (outcome variables)}
z1 <- mod_theta$theta1/mod_theta$setheta1
p1 <- (1 - pnorm(abs(z1), 0, 1)) * 2

z2 <- mod_theta$theta2/mod_theta$setheta2
p2 <- (1 - pnorm(abs(z2), 0, 1)) * 2
```
 
```{r storing thresholds and P values for thresholds (outcome variables)}

mod_theta <- data.frame(thresholds1 = c("VLOW|LOW",
                                     "LOW|MODERATE",
                                     "MODERATE|HIGH",
                                     "HIGH|VHIGH"),
                        theta1 = mod_bivariate$theta$r1A_STRESS, 
                        setheta1 = mod_bivariate$setheta$r1A_STRESS,
                        p1 = p1,
                        
                        
                        thresholds2 = c("NOT IMPORTANT|SLIGHTLY IMPORTANT",
                                     "SLIGHTLY IMPORTANT|MODERATELY IMPORTANT",
                                     "MODERATELY IMPORTANT|IMPORTANT",
                                     "IMPORTANT|VERY IMPORTANT"),
                        theta2 = mod_bivariate$theta$r1GA_IMPSTRESS,
                        setheta2 = mod_bivariate$setheta$r1GA_IMPSTRESS,
                        p2 = p2)


```


```{r Storing coefficients for exploratory variables}
(mod_beta <- coef(summary(mod_bivariate)))


# Define new variable names
new_row_names <- c("r0D_AGEAdults 2", "r0D_AGESeniors 1", "r0J_INCOMEMiddle Income 1", "r0J_INCOMEMiddle Income 2", "r0J_INCOMEHigh Income 1", "r0P_MODE1Private Transportation 1", "r0P_MODE1Public Transportation 1", "r0G_EDUCATIONHigher Education 2", "r0H_OCCUPATIONWorking 1", "r0H_OCCUPATIONStudent 1")


# Assign new names to the variables
rownames(mod_beta) <- new_row_names

# Print the modified coefficient summary
print(mod_beta)
```

```{r Calculating Z and P values for exploratory variables}

# data_ <- as.data.frame(t(mod_bivariate))
# mod_bivariate[["beta"]][["r0D_AGEAdults 1"]]
# 
# z_Seniors <- 'beta'$'r0D_AGESeniors 1'/'sebeta'$'r0D_AGESeniors 1'
# p_Seniors <- (1 - pnorm(abs(z_Seniors), 0, 1)) * 2





#for equation 1
z_Seniors <- mod_bivariate$beta[2]/mod_bivariate$sebeta[2]
p_Seniors <- (1 - pnorm(abs(z_Seniors), 0, 1)) * 2

z_Middle_Income_1 <- mod_bivariate$beta[3]/mod_bivariate$sebeta[3]
p_Middle_Income_1 <- (1 - pnorm(abs(z_Middle_Income_1), 0, 1)) * 2

z_High_Income <-mod_bivariate$beta[5]/mod_bivariate$sebeta[5]
p_High_Income <- (1 - pnorm(abs(z_High_Income), 0, 1)) * 2

z_Private_T <- mod_bivariate2$beta[6]/mod_bivariate2$sebeta[6]
p_Private_T <- (1 - pnorm(abs(z_Private_T), 0, 1)) * 2

z_Public_T <- mod_bivariate$beta[7]/mod_bivariate$sebeta[7]
p_Public_T<- (1 - pnorm(abs(z_Public_T), 0, 1)) * 2

z1_Working <- mod_bivariate$beta[6]/mod_bivariate$sebeta[6]
p_Working <- (1 - pnorm(abs(z1_Working), 0, 1)) * 2

z_Student <- mod_bivariate$beta[10]/mod_bivariate$sebeta[10] 
p_Student <- (1 - pnorm(abs(z_Student), 0, 1)) * 2



#for equation 2
z_Adults <- mod_bivariate$beta[1]/mod_bivariate$sebeta[1]
p_Adults <- (1 - pnorm(abs(z_Adults), 0, 1)) * 2

z_Middle_Income_2 <-mod_bivariate$beta[4]/mod_bivariate$sebeta[4] 
p_Middle_Income_2 <- (1 - pnorm(abs(z_Middle_Income_2), 0, 1)) * 2

z_Higher_Education <- mod_bivariate$beta[8]/mod_bivariate$sebeta[8]
p_Higher_Education <- (1 - pnorm(abs(z_Higher_Education), 0, 1)) * 2
```


```{r storing estimates and P values for exploratory variables}
mod_beta <- data.frame(Equation1 = c(NA,
                                      "r0D_AGESeniors 1",
                               "r0J_INCOMEMiddle Income 1",
                               NA,
                               "r0J_INCOMEHigh Income 1", 
                               "r0P_MODE1Private Transportation 1",
                               "r0P_MODE1Public Transportation 1",
                               NA,
                               "r0H_OCCUPATIONWorking 1", 
                               "r0H_OCCUPATIONStudent 1"),
                        beta_equation1 = c( NA,
                                            mod_bivariate[["beta"]][["r0D_AGEAdults 1"]],
                                            mod_bivariate[["beta"]][["r0J_INCOMEMiddle Income 1"]],
                                            NA,
                                            mod_bivariate[["beta"]][["r0J_INCOMEHigh Income 1"]],
                                            mod_bivariate[["beta"]][["r0P_MODE1Private Transportation 1"]],
                                            mod_bivariate[["beta"]][["r0P_MODE1Public Transportation 1"]],
                                            NA,
                                            mod_bivariate[["beta"]][["r0H_OCCUPATIONWorking 1"]],
                                            mod_bivariate[["beta"]][["r0H_OCCUPATIONStudent 1"]]),
                          
                          
                          
                        se_equation1 = c( NA,
                                            mod_bivariate[["sebeta"]][["r0D_AGEAdults 1"]],
                                            mod_bivariate[["sebeta"]][["r0J_INCOMEMiddle Income 1"]],
                                            NA,
                                            mod_bivariate[["sebeta"]][["r0J_INCOMEHigh Income 1"]],
                                            mod_bivariate[["sebeta"]][["r0P_MODE1Private Transportation 1"]],
                                            mod_bivariate[["sebeta"]][["r0P_MODE1Public Transportation 1"]],
                                            NA,
                                            mod_bivariate[["sebeta"]][["r0H_OCCUPATIONWorking 1"]],
                                            mod_bivariate[["sebeta"]][["r0H_OCCUPATIONStudent 1"]]),
                       
                       p_values_E1 = c (NA, p_Seniors, p_Middle_Income_1, NA,  p_High_Income, p_Private_T, p_Public_T, NA, p_Working,  
                                        p_Student),
                        
                        #  NA,
                        # p_Seniors = p_Seniors,
                        # p_Middle_Income_1 = p_Middle_Income_1,
                        # NA,
                        # p_High_Income = p_High_Income,
                        # p_Private_T = p_Private_T,
                        # p_Public_T = p_Public_T,
                        # NA,
                        # p_Working = p_Working,
                        # p_Student = p_Student,
                
                        
                        
                        
                               
                        Equation2 = c("r0D_AGEAdults 1",
                                      NA,
                                      NA,
                                "r0J_INCOMEMiddle Income 2",
                                NA,
                                NA,
                                NA,
                                "r0G_EDUCATIONHigher Education 1",
                                NA,
                                NA),
                                
                              
                        beta_equation2 = c (mod_bivariate[["beta"]][["r0D_AGEAdults 1"]],
                                            NA,
                                            NA,
                                            mod_bivariate[["beta"]][["r0J_INCOMEMiddle Income 2"]],
                                            NA,
                                            NA,
                                            NA,
                                            mod_bivariate[["beta"]][["r0G_EDUCATIONHigher Education 1"]],
                                            NA,
                                            NA),
                        
                        
                        se_equation2 = c( mod_bivariate[["sebeta"]][["r0D_AGEAdults 1"]],
                                         NA,
                                         NA,
                                         mod_bivariate[["sebeta"]][["r0J_INCOMEMiddle Income 2"]],
                                         NA,
                                         NA,
                                         NA,
                                         mod_bivariate[["sebeta"]][["r0G_EDUCATIONHigher Education 1"]],
                                         NA,
                                         NA),
                                        
                        # p_Adults = p_Adults,
                        # NA,
                        # NA,
                        # p_Middle_Income_2 = p_Middle_Income_2,
                        # NA,
                        # NA,
                        # NA,
                        # p_Higher_Education = p_Higher_Education,
                        # NA,
                        # NA)
                         
                       p_values_E2 = c (p_Adults, NA, NA, p_Middle_Income_2, NA, NA, NA, p_Higher_Education, NA, NA))
  
```




```{r assemble-table-with-results, include=FALSE}
# Join the coefficient estimates and p-values to the table
mod_table <- mod_table |>
  # Estimates of equation 1
  left_join(mod_beta,
            by = c("e1_var" = "variable")) |>
  rename(e1_beta = beta) |>
  # Estimates of equation 2
  left_join(mod_beta,
            by = c("e2_var" = "variable")) |>
  rename(e2_beta = beta) |>
  # p-values of equation 1
  left_join(mod_beta,
            by = c("e1_var" = "variable")) |>
  rename(e1_p = p) |>
  # p-values of equation 2
  left_join(mod_beta,
            by = c("e2_var" = "variable")) |>
  rename(e2_p = p) |>
  # Select columns for table
  dplyr::select(variable,
                e1_beta,
                e1_p,
                e2_beta,
                e2_p)
```


```{r collect-coefficients-and-thresholds, include=FALSE}
# Check column names
colnames(mod_beta)
colnames(mod_theta)

# Rename columns to match
colnames(mod_theta) <- colnames(mod_beta)

# Join table of coefficient estimates and thresholds
mod_table1 <- rbind(mod_beta, mod_theta)

```


```{r model-results, echo=FALSE}
mod_table1 |>  
  mutate(across(.cols = ends_with("beta"),
                ~ ifelse(is.na(.),
                         "-",
                         paste(round(., 4)))),
         across(.cols = ends_with("p"),
                ~ ifelse(is.na(.),
                         "-",
                         ifelse(. >= 0.0001,
                                paste(round(., 4)),
                                "<0.0001")))) |>
  kable("latex",
        booktabs = T,
        col.names = c("Variable", "Estimate", "p-value", "Estimate", "p-value"),
        align = c("l", "c", "c", "c", "c"),
        caption = "\\label{tab:model-results}Bivariate ordered model results: walking frequency before and during COVID-19") |>
  kable_styling(font_size = 8,
                latex_options = c("striped")) #|>
# pack_rows(group_label = "Socio-economic and demographic variables",
#           start_row = 1,
#           end_row = 6) |>
# pack_rows(group_label = "Regular mode of transportation before COVID-19",
#           start_row = 5,
#           end_row = 9) |>
# pack_rows(group_label = "Attitudinal variables",
#           start_row = 10,
#           end_row = 11) |>
#pack_rows(group_label = "Thresholds",
#          start_row = 27,
#          end_row = 31)
```

```{r displaying the result}

library(knitr)
library(kableExtra)
 

mod_table1 %>%
  mutate(across(.cols = ends_with("mod_beta"),
                ~ ifelse(is.na(.), "-", paste(round(., 4)))),
         across(.cols = ends_with("mod_theta"),
                ~ ifelse(is.na(.),
                         "-",
                         ifelse(. >= 0.0001,
                                paste(round(., 4)),
                                "<0.0001")))) %>%
  kable(format = "latex",
        booktabs = TRUE,
        col.names = c("Equation1", "beta_equation1", "se_equation1", "p_values_E1", "Equation2", "beta_equation2", "se_equation2", "p_values_E2"),
        align = c("l", "c", "c", "c", "c"),
        caption = "\\label{tab:model-results}Bivariate ordinal model results: Stress level and importance of experienced stress during commute") %>%
  kable_styling(font_size = 8,
                latex_options = c("striped"))


```

<!-- Previous visualizations
```{r}
# Load the required library
library(ggplot2)

# Create a subset of the dataset containing only the relevant variables
subset_data <- df[, c("r0D_AGE", "r0P_MODE1")]

# Create a stacked bar plot
ggplot(subset_data, aes(x = factor(r0D_AGE, levels = c("Youth","Adults", "Seniors")), fill = r0P_MODE1)) +
  geom_bar(position = "fill") +
  labs(x = "Age Group", y = "Proportion", fill = "Transportation Mode") +
  scale_fill_discrete(name = "Transportation Mode", labels = c("Private Transportation", "Public Transportation", "Active Transportation")) +
  ggtitle("Transportation Mode by Age Group") +
  theme_minimal()

```
Mode of transportation and age groups: Age groups look similar in private mode but youth are more willing to use active than two others. Adults and seniors willing to use public more than youth.
```{r}
# Load the required library
library(ggplot2)

# Create a subset of the dataset containing only the relevant variables
subset_data <- df[, c("r0D_AGE", "r1A_STRESS")] %>%
drop_na(r0D_AGE, r1A_STRESS)

# Create a stacked bar plot
ggplot(subset_data, aes(x = factor(r0D_AGE, levels = c("Youth","Adults", "Seniors")), fill = r1A_STRESS)) +
  geom_bar(position = "fill") +
  labs(x = "Age Group", y = "Proportion", fill = "Experienced stress level") +
  scale_fill_discrete(name = "Experienced stress level", labels = c("LOW", "MODERATE", "HIGH", "VHIGH")) +
  ggtitle("Experienced stress level by Age Group") +
  theme_minimal()

```
Experienced stress and age groups: Youth and Adults are more likely to experience VHIGH level stress than seniors
```{r}
# Load the required library
library(ggplot2)

# Create a subset of the dataset containing only the relevant variables
subset_data <- df[, c("r0P_MODE1", "r1A_STRESS")] %>%
drop_na(r0P_MODE1, r1A_STRESS)

# Create a stacked bar plot
ggplot(subset_data, aes(x = factor(r0P_MODE1, levels = c("Private Transportation", "Public Transportation", "Active Transportation")), fill = r1A_STRESS)) +
  geom_bar(position = "fill") +
  labs(x = "Mode", y = "Proportion", fill = "Experienced stress level") +
  scale_fill_discrete(name = "Experienced stress level", labels = c("LOW", "MODERATE", "HIGH", "VHIGH")) +
  ggtitle("Transportation Mode by Stress Group") +
  theme_minimal()

```
Mode of transportation and experienced stress: Kind of similar except for VHIGH in private (more used by youth) and public (have been used more by adults and seniors) modes than active.

Common thing among these three bar charts would be adults for using more public transportation.
-->

Regarding analysis during our bivariate ordinal mode, we reclassified different modes of transportation into three general categories of active transportation (walking and cycling), public transportation (taxi, collective, metro and bus) and private transportation (car and motorcycle). 

As can be seen in the bar chart, respondent mostly use public means of transportation with almost 285 out of 451 respondents as their main choice for commuting. Recent studies also have shown that using public transport has formed almost 50% of motorized trips as a salient example of the popularity of public transportation in Santiago, Chile over the recent years [@pezoa2023estimation]. Regarding this popularity and to make it more productive, it has been mentioned that public transport in Santiago as a modern and integrated system is called Transantiago referring to a sustainable public transport system for Santiago. By aggregating services such as bus and metro and keeping the fare expenses with a low range, passengers can benefit from an efficient system of transport which was unsuccessful [@munoz2008transantiago] and needs some consideration to work as planners expected.
Following this, second popular mode of transportation can be assigned to private mode with around 25% of the whole population or 120 persons. Some users also were interested in eco-friendly modes like walking and cycling with slightly more than 40 people and 10 % of the whole respondents.

```{r Modes of transport}


ggplot(df, aes(x=as.factor(r0P_MODE1), fill=as.factor(r0P_MODE1) )) + 
  geom_bar( ) +
  scale_fill_hue(c = 50) +
  theme(legend.position="none")+
    scale_y_continuous(breaks = seq(0, 451, 20))

```

To have a better understanding of the interaction among the key variables of modes of transportation, age groups, income levels, experienced levels of stress and the importance assigned to the stress, some mosaic plots were adopted. 

```{r}
req <- substitute(require(x, character.only = TRUE))
libs<-c("sjPlot")
sapply(libs, function(x) eval(req) || {install.packages(x); eval(req)})
```
Mode, age and income are the most crucial variables in our study. As it has been studied in previous work, different modes of commuting can be effective on people's feeling and in general the subjective well-being referring to the importance of focusing on individuals' travel experiences such as stress [@smith2017commute].Income also is one of the significant variables as it determine people's options for commuting like if they are low income workers they tend to use more public transportation in Santiago, Chile [@gomez2023urban] which could be important from stress of commute's perspective that can be seen mostly among public transport users in this study. Another influential variable on people's experience of commuting is age which sometimes makes commuting difficult for older people specially when they choose walking according to the folks in La Cisterna district to the city center in Santiago [@tironi2016affects].   

So we are analyzing these items in terms of their effect on each other.



```{r display level of Experienced Stress by the importance of stress}
sjPlot::tab_xtab(var.row = df$r1A_STRESS, var.col = df$r1GA_IMPSTRESS, title = "Table Title", show.row.prc = TRUE)
```

As  can be seen in the cross tabulation of experienced stress level and importance of that while commute, almost half of those who assigned very low, low and moderate level to experienced stress felt it is very important to them. Also those who felt stress at low and moderate levels formed about a quarter of the whole population each with respectively 104 and 141 people .When it comes to those with high and very high levels of feeling stress, around two third of these people assigned very high level of importance to their feeling. 

Generally, at lower levels it has been mostly less important for people and the higher levels of stress, the more importance that people assign.

Results can be visible also using the following mosaic plot.
 
```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r1GA_IMPSTRESS, r1A_STRESS), fill=r1GA_IMPSTRESS)) + 
  labs(title = "Importance of Stress by Experienced stress level")

```


```{r display level of Experienced Stress by transportation modes}
sjPlot::tab_xtab(var.row = df$r0P_MODE1, var.col = df$r1A_STRESS, title = "Table Title", show.row.prc = TRUE)
```

According to the cross tabulation of modes of transport by feeling of stress, almost a large group of active mode users experienced very low and moderate level of stress while commuting. It can be seen that most people who use private and public transport felt low to high level of stress in these groups. Public transport is the most popular means of transportation in this study with almost half of people. 

So as a whole fact private and public transports cause relatively higher levels of stress compared to active mode. 


```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r0P_MODE1, r1A_STRESS), fill=r0P_MODE1)) + 
  labs(title = "Transportation Mode by Experienced Stress")

```



```{r display level of importance of stress by transportation modes }
sjPlot::tab_xtab(var.row = df$r0P_MODE1, var.col = df$r1GA_IMPSTRESS, title = "Table Title", show.row.prc = TRUE)
```


Results of the cross tabulation of importance of stress level and different means of transportation indicated that a remarkable number of people using active mode have assigned important and very important level to the stress feeling they have experienced. Similarly, public transport users also assigned the same level to the stress feeling but this group takes the largest number of people with 279 out of 451 people. As the second group in terms of popularity, private transport users also mostly have given important and very important level of stress feeling on commute with almost 25 % and 60% of them. Interesting point is that among all kinds of transports, the largest group of people assigned very important level to the experienced stress were public users with 235 people out of 451 respondents.

Totally it is important or very important for the users of all modes of transport to mention their stress level and anxiety feelings while commuting.



```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r0P_MODE1, r1GA_IMPSTRESS), fill=r0P_MODE1)) + 
  labs(title = "Transportation Mode by Importance of Stress")

```



```{r display level of importance of stress by Income level }
sjPlot::tab_xtab(var.row = df$r0J_INCOME, var.col = df$r1GA_IMPSTRESS, title = "Table Title", show.row.prc = TRUE)
```

Based on the cross tabulation of importance of stress by income level, it can be seen that among all three groups of income around 80 % of population put important and very important levels on feeling stress when commuting. It is also been indicated that middle income and high-income population formed almost 70 % of the whole population altogether.

It is interesting to note that a large number of respondents in all income groups assigned important and very important level to experienced stress during trips.
 

```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r1GA_IMPSTRESS, r0J_INCOME), fill=r1GA_IMPSTRESS)) + 
  labs(title = "Importance of Stress by Income level")

```


```{r display level of Experienced Stress by Income level }
sjPlot::tab_xtab(var.row = df$r0J_INCOME, var.col = df$r1A_STRESS, title = "Table Title", show.row.prc = TRUE)
```

According to the cross tabulation of feeling of stress by income level it has been indicated that mostly low income group felt moderate about stress and other low income respondents were dispersed in the different feeling levels. On the contrary, a huge number of people belong to middle level of income (almost 60 %) felt low and moderate levels of experiencing stress and less than 20 % of them felt high and very high levels of stress of commuting. This can be considered in a similar way for high income level people as well because more than 80% of them experienced low to high levels of stress. 

Totally the larger part of population including middle and high income felt less challenging ranges of stress in comparison to low income levels.
 


```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r1A_STRESS, r0J_INCOME), fill=r1A_STRESS)) + 
  labs(title = "Experienced stress level by Income level")

```


```{r display level of Mode by Income level }
sjPlot::tab_xtab(var.row = df$r0J_INCOME, var.col = df$r0P_MODE1, title = "Table Title", show.row.prc = TRUE)
```

According to cross tabulation of mode choice by income level, almost 80% of lower level income people and 65% of middle income levels mostly use public transport which has been indicated to be stressful. High income respondents where interested in using public and private in a relatively similar way with about 40% and 50% respectively.  
```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r0P_MODE1, r0J_INCOME), fill=r0P_MODE1)) + 
  labs(title = "Experienced stress level by Income level")

```



```{r display level of Importance of Stress by Age groups }
sjPlot::tab_xtab(var.row = df$r0D_AGE, var.col = df$r1GA_IMPSTRESS, title = "Table Title", show.row.prc = TRUE)
```

As can be seen in the cross tabulation of importance of stress by age groups, most of population includes youth and adults with almost 30% and 60% of the whole people. Seniors also has incorporated around 10% of people which is the least populated group. It can be seen that there is a common trend among all groups that around 70% of each group put important and very important levels on experiencing stress while commute. Surprisingly, there were almost two third of adults who assigned very important level of stress when commuting which is the highest record among others.

In general the given importance range by each age group were almost the same but adults went drastically up to the highest number of people who assigned very important level to the commuting stress.


```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r1GA_IMPSTRESS, r0D_AGE), fill=r1GA_IMPSTRESS)) + 
  labs(title = "Importance of Stress by Age Group")

```

```{r display level of Experienced stress by Age groups}
sjPlot::tab_xtab(var.row = df$r0D_AGE, var.col = df$r1A_STRESS, title = "Table Title", show.row.prc = TRUE)
```

According to the cross tabulation of experiencing stress by age groups, it can be seen that for youth and adults as the larger part of respondents there is a kind of similar dispersion of feeling stress in the range of low to high level with about more than 70% of people in each age group. This trend can not be responsive for seniors as they felt very low to very high levels of stress with an almost same portions for each level except high level of experiencing stress. It can be notable that seniors went with almost the same per cents for each level but it went down to only about 10% of seniors when they felt high level of stress while commuting.

Totally adults and youth experienced higher levels of stress than their Senior counterparts.

```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r1A_STRESS, r0D_AGE), fill=r1A_STRESS)) + 
  labs(title = "Experienced stress level by Age Group")

```

```{r}
df |> 
  filter(!is.na(r1A_STRESS), !is.na(r1GA_IMPSTRESS), !is.na(r0P_MODE1)) |>
ggplot() + 
  geom_mosaic(aes(x = product(r1A_STRESS,r1GA_IMPSTRESS), fill= r1A_STRESS)) + 
  labs(title = "Experienced stress level by Age Group") +
  theme(axis.text.x = element_text(angle = 90, size = 5)) + 
  facet_wrap(~r0P_MODE1)

```





```{r display Age groupd by transportation modes}
sjPlot::tab_xtab(var.row = df$r0D_AGE, var.col = df$r0P_MODE1, title = "Table Title", show.row.prc = TRUE)
```

It can be seen from the cross tabulation of modes of transport by age groups that youth as the most populated age group were interested in using public transport with more than 70% of them. Following this, adults as the second most populated age group were interested in using both private and public transportation with almost the same share (60% of adults for each mode). Apart from this, seniors were mostly interested in public transport with slightly less than 60%. Moreover, second popular mode among seniors was private transport with 35% of senior respondents.

So youth respondents tend to use public mode, adults and seniors interested mostly in using both public and private modes.


```{r}
ggplot(data = df) +
  geom_mosaic(aes(x = product(r0D_AGE, r0P_MODE1), fill=r0D_AGE)) + 
  labs(title = "Transportation Mode by Age group")

```

1-Experienced stress and importance of stress
Generally, at lower levels it has been less important for people and the higher levels of stress, the more importance that people assign.

                          
                                       
2-1-Experienced stress and mode
So as a whole fact private and public transports cause relatively higher levels of stress compared to active mode. 

public and private = higher stress = higher importance

2-2-Importance of stress and mode
Totally it is important or very important for the users of all modes of transport to mention their stress level and anxiety feelings while commuting.





3-1-Importance of stress and income 
It is interesting to note that a large number of respondents in all income groups assigned important and very important level to experienced stress during trips


3-2-Experienced stress and income
Totally the larger part of population including middle and high income felt less challenging ranges of stress in comparison to low income levels.



3-3-Mode and income

Low income and middle income using more public. High income use both private and public with relatively similar share




4-1-Importance of stress and age
In general the given importance range by each age group were almost the same but adults went drastically up to the highest number of people who assigned very important level to the commuting stress.


4-2-Experienced stress and age
Totally adults and youth experienced higher levels of stress than their Senior counterparts.




5-1-Age and mode
So youth respondents tend to use public mode, adults and seniors interested mostly in using both public and private modes.


In general those adults who use public transport and belong to low-income level are the most group at the exposure of feeling stress while they mention it is also very important to them. 

# Conclusion

This research has been conducted to investigate effect of commuting stress on choosing mode of active and motorized means of transportation and the corresponding coping strategies. Recently, there has been increased attention in how these consequences are felt by travelers who utilize various modes of transportation, particularly active ones.

Measuring stress entails an individual self-evaluation of several variables when travelling – preferably while commuting. There is a link between stress and commuting in the literature, which takes into account factors such as commute time [@evans2006rail], control and predictability [@gottholmseder2009stress], as already has been found. The component of stress level has a negative connotation in the context of this research (as no one would prefer to be stressed) and investigates the internal mental condition of the person who travels – based on her commuting journeys. 

In this research we aimed at adding psychological repercussions on travelers by investigating stress levels among users of active and motorized modes of transportation in this research. Furthermore, we look into the significance that travelers place on their stress levels. This allows us to investigate the concept of "limited horizons," or how those who are less adaptable normalize subpar experiences. The current study on commuting stress addressed the following questions: 
-How crucial is it for respondents  to make a journey while feeling a sense of stress and to what scale they are going to evaluate this feeling?
-What elements go into determining how stressed active and motorized travelers are when traveling?
-Can commuters think of normalization and learned helplessness as coping mechanisms? If yes, which category of people.

<!-- study limitations -->

<!--chapter:end:02-chap2.Rmd-->

# Conclusion {-}

If we don't want Conclusion to have a chapter number next to it, we can add the `{-}` attribute.

**More info**

And here's some other random info: the first paragraph after a chapter title or section head _shouldn't be_ indented, because indents are to tell the reader that you're starting a new paragraph. Since that's obvious after a chapter or section title, proper typesetting doesn't add an indent there.


<!--chapter:end:03-conclusion.Rmd-->

<!--
The bib chunk below must go last in this document according to how R Markdown renders.  More info is at http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
-->

\backmatter

<!-- 
If you'd like to change the name of the bibliography to something else,
delete "References" and replace it.
-->

# References {-}
<!--
This manually sets the header for this unnumbered chapter.
-->
\markboth{References}{References}
<!--
To remove the indentation of the first entry.
-->
\noindent

<!--
To create a hanging indent and spacing between entries.  These three lines may need to be removed for styles that don't require the hanging indent.
-->

\setlength{\parindent}{-0.20in}
\setlength{\leftskip}{0.20in}
\setlength{\parskip}{8pt}


<!--
This is just for testing with more citations for the bibliography at the end.  Add other entries into the list here if you'd like them to appear in the bibliography even if they weren't explicitly cited in the document.
-->

---
nocite: | 
  @angel2000, @angel2001, @angel2002a
...

<!--chapter:end:99-references.Rmd-->

